---
title: "Course Overview"
---

```{r}
#| label: setup
#| include: false

library(ggplot2)

knitr::opts_chunk$set(
  fig.width = 7,
  fig.height = 4.6,
  out.width = "100%",
  dpi = 160,
  fig.retina = 2,
  comment = NA
)

theme_set(
  theme_minimal(base_size = 13) +
    theme(
      plot.title.position = "plot",
      panel.grid.minor = element_blank()
    )
)

set.seed(1234)
```

## Welcome

This Quarto book is designed as a single, well-structured version of a Network Science course.

Each chapter follows the same pattern:

1.  **Theory and intuition**
2.  **Worked examples in R**
3.  **Exercises and coding tasks**

The same source files produce both:

-   an HTML course website
-   a PDF course handout

## Required packages

Install the packages below once:

```{r}
#| eval: false
install.packages(c(
  "igraph", "dplyr", "ggplot2", "tidyr", "purrr", "plotly", "tinytex", "quarto", "ggraph"
))
tinytex::install_tinytex()
```

## Current chapters

The current book contains these chapters:

- [Random Graphs](chapters/03-random-graphs.qmd)
- [Random Graph Evolution](chapters/04-random-graph-evolution.qmd)
- [Small-World Graphs](chapters/05-small-world-graphs.qmd)
- [Scale-free Graphs](chapters/06-scale-free-graphs.qmd)


