---
title: "Course Overview"
---

```{r}
#| label: setup
#| include: false

library(ggplot2)

knitr::opts_chunk$set(
  fig.width = 7,
  fig.height = 4.6,
  out.width = "100%",
  dpi = 160,
  fig.retina = 2,
  comment = NA
)

theme_set(
  theme_minimal(base_size = 13) +
    theme(
      plot.title.position = "plot",
      panel.grid.minor = element_blank()
    )
)

set.seed(1234)
```

## Welcome

This Quarto book is designed as a single, well-structured version of a Network Science course.

Each chapter follows the same pattern:

1. **Theory and intuition**
2. **Worked examples in R**
3. **Exercises and coding tasks**

The same source files produce both:

- an HTML course website
- a PDF course handout

## Required packages

Install the packages below once:

```{r}
#| eval: false
install.packages(c(
  "igraph", "dplyr", "ggplot2", "tidyr", "purrr", "plotly", "tinytex", "quarto"
))
tinytex::install_tinytex()
```

## Design principles

This template is built around a few teaching priorities:

- explanations should be readable and beginner-friendly
- formulas should be introduced with intuition, not only stated
- code should be clean enough to use directly in class
- plots should be visually clear and consistent
- HTML pages may contain interactivity, while PDF remains clean and stable

## Current chapters

The current book contains these chapters:

1. **Random Graphs**
2. **The Evolution of Random Graphs**

The first chapter introduces random graph ensembles and their basic structural properties. The second chapter studies how global connectivity emerges as a random graph evolves.