---
title: "Random Graph Evolution — Solutions"
subtitle: "Corresponds to `chapters/04-random-graph-evolution.qmd`"
---

```{r}
#| label: solution-setup
#| include: false

setup_candidates <- c("../../R/book-setup.R", "../R/book-setup.R", "R/book-setup.R")
setup_file <- setup_candidates[file.exists(setup_candidates)][1]

if (is.na(setup_file)) {
  stop("Could not locate R/book-setup.R", call. = FALSE)
}

source(setup_file, local = knitr::knit_global())
ns_book_setup(preset = "evolution", seed = 42)
```

## Exercise 1. Components, partitions, and the giant component criterion

**Part 1.** *Proof that components partition $V$.*

Define a relation $\sim$ on $V$ by $u \sim v$ if there exists a path from $u$ to $v$ in $G$. This is an equivalence relation: reflexivity holds because the empty path connects $v$ to itself; symmetry holds because paths are undirected; transitivity holds because two paths can be concatenated. The equivalence classes of $\sim$ are exactly the connected components: they are connected (any two vertices in the same class are path-connected by definition) and maximal (adding a vertex from a different class would require a cross-class path, contradicting the class partition). Since equivalence classes partition any set, the components partition $V$. $\square$

**Part 2.** For $P_N$ (path on $N$ vertices), the graph is already connected, so $C_{\max} = P_N$ and $|C_{\max}|/N = N/N = 1 \to 1$. The path graph trivially contains a "giant component" in the sense that $c = 1 > 0$. However, this is uninteresting — the fraction is $1$ for all $N$, not emerging through a random process. The definition is satisfied, but $P_N$ is a degenerate case: it is deterministically connected.

**Part 3.** For $\lfloor N/2\rfloor$ disjoint edges ($K_2$ matchings), each component has size $2$ (or $1$ if $N$ is odd). So $|C_{\max}| = 2$ for all $N \geq 2$, giving $|C_{\max}|/N = 2/N \to 0$. This sequence does **not** contain a giant component — the largest component is microscopic, not macroscopic.

**Part 4.** One construction: take $\lfloor N/3\rfloor$ cliques $K_3$. Then the largest component has size $3$, so $|C_{\max}|/N \to 0$. To get $S = 1/3$, we need a deterministic construction where exactly $N/3$ vertices are in one component and the rest in components of size $o(N)$. For example: one clique $K_{\lfloor N/3\rfloor}$ and $\lceil 2N/3\rceil$ isolated vertices. Then $|C_{\max}|/N \to 1/3$. This satisfies $c = 1/3 > 0$, so it is a giant component by definition.

For the ER model: solve $1/3 = 1 - e^{-\langle k\rangle/3}$, i.e., $e^{-\langle k\rangle/3} = 2/3$, so $\langle k\rangle = -3\ln(2/3) = 3\ln(3/2) \approx 1.216$.

```{r}
#| label: ex1-verify
#| echo: true

# Verify numerically
f <- function(S, k) 1 - exp(-k * S)
k_target <- -3 * log(2/3)
cat("k for S = 1/3:", round(k_target, 4), "\n")
cat("Check: f(1/3, k) =", round(f(1/3, k_target), 6), "\n")
```

---

## Exercise 2. The self-consistency equation: analysis and phase transition

**Part 1.** $f(0) = 1 - e^0 = 0$, confirming $S = 0$ is always a fixed point. For any $S^* > 0$: $f(S^*) = 1 - e^{-\langle k\rangle S^*} < 1$ since the exponential is positive. So $S^* < 1$ for any finite $\langle k\rangle$. $\square$

**Part 2.** Define $g(S) = f(S) - S = 1 - e^{-\langle k\rangle S} - S$.

- $g(0) = 0$.
- $g'(S) = \langle k\rangle e^{-\langle k\rangle S} - 1$.
- At $S = 0$: $g'(0) = \langle k\rangle - 1$.

If $\langle k\rangle \leq 1$: $g'(0) \leq 0$. Since $g''(S) = -\langle k\rangle^2 e^{-\langle k\rangle S} < 0$, $g$ is strictly concave. Combined with $g(0) = 0$ and $g'(0) \leq 0$, the function $g$ is non-positive for all $S > 0$, meaning $f(S) \leq S$ for all $S > 0$. Therefore no positive fixed point exists.

If $\langle k\rangle > 1$: $g'(0) > 0$, so $g$ is initially increasing. Since $g$ is concave and $g(S) \to -\infty$ as $S \to \infty$ (dominated by $-S$), the intermediate value theorem guarantees exactly one positive root $S^* > 0$. $\square$

**Part 3.**

```{r}
#| label: ex2-fixed-points
#| echo: true
#| fig-cap: "Theoretical giant component fraction S* from the self-consistency equation, with simulation estimates overlaid."

solve_S <- function(k_mean, tol=1e-10) {
  if (k_mean <= 1) return(0)
  S <- 0.5
  for (i in 1:10000) {
    S_new <- 1 - exp(-k_mean * S)
    if (abs(S_new - S) < tol) return(S_new)
    S <- S_new
  }
  S
}

k_vals <- seq(0.01, 5, length.out=300)
S_vals <- sapply(k_vals, solve_S)

# Spot checks
for (k in c(1.5, 2, 3, 5)) {
  S <- solve_S(k)
  cat(sprintf("k = %.1f: S* = %.4f, check: 1 - exp(-kS) = %.4f\n",
              k, S, 1 - exp(-k*S)))
}

# Simulation estimates at selected k values
set.seed(42)
n_sim <- 500
k_sim <- seq(0.5, 5, by=0.5)
sim_S <- sapply(k_sim, function(k) {
  p <- k / (n_sim - 1)
  mean(sapply(1:100, function(i) {
    g <- sample_gnp(n_sim, p, directed=FALSE, loops=FALSE)
    max(components(g)$csize) / n_sim
  }))
})

theory_df <- tibble(k=k_vals, S=S_vals)
sim_df    <- tibble(k=k_sim, S=sim_S)

ggplot(theory_df, aes(x=k, y=S)) +
  geom_line(color=course_cols$teal, linewidth=1.2) +
  geom_point(data=sim_df, aes(x=k, y=S), color=course_cols$gold, size=2.5) +
  geom_vline(xintercept=1, color=course_cols$coral, linetype=2) +
  labs(title="Giant component fraction: theory vs simulation",
       x=expression(langle*k*rangle), y="S*")
```

**Part 4.** A first-order transition would mean $S$ jumps discontinuously from $0$ to some $S_0 > 0$ at $\langle k\rangle = 1$. Physically, this would mean the giant component appears instantaneously with a finite fraction of all vertices — an abrupt onset with no intermediate stage. Instead, the second-order (continuous) transition means that just above $\langle k\rangle = 1$, only a tiny fraction of vertices join the giant component, growing linearly as $\langle k\rangle$ increases. This is analogous to a continuous phase transition in physics (e.g., the order parameter growing continuously from zero at the critical temperature in a ferromagnet).

---

## Exercise 3. Near-threshold expansion

**Part 1.** Expand $e^{-\langle k\rangle S}$ to third order:

$$
e^{-\langle k\rangle S} = 1 - \langle k\rangle S + \frac{\langle k\rangle^2 S^2}{2} - \frac{\langle k\rangle^3 S^3}{6} + O(S^4).
$$

So:

$$
S = 1 - e^{-\langle k\rangle S} = \langle k\rangle S - \frac{\langle k\rangle^2 S^2}{2} + \frac{\langle k\rangle^3 S^3}{6} + O(S^4).
$$

Rearranging: $S(1 - \langle k\rangle) + \frac{\langle k\rangle^2 S^2}{2} - \frac{\langle k\rangle^3 S^3}{6} = 0$. Dividing by $S \neq 0$:

$$
(1 - \langle k\rangle) + \frac{\langle k\rangle^2 S}{2} - \frac{\langle k\rangle^3 S^2}{6} = 0.
$$

Writing $\langle k\rangle = 1 + \varepsilon$ and $S = a_1\varepsilon + a_2\varepsilon^2 + O(\varepsilon^3)$, the leading-order term gives $-\varepsilon + \frac{S}{2} \approx 0$, so $S \approx 2\varepsilon$. The correction term $-\frac{\langle k\rangle^3 S^2}{6} \approx -\frac{\varepsilon^2 \cdot 4\varepsilon^2}{6}$ is $O(\varepsilon^2)$ relative to $S \sim \varepsilon$, confirming the linear approximation is valid to leading order.

**Part 2.** Substituting $\langle k\rangle = 1+\varepsilon$ and $S = a_1\varepsilon + a_2\varepsilon^2$ into the equation:

$(−\varepsilon) + \frac{(1+\varepsilon)^2}{2}(a_1\varepsilon + a_2\varepsilon^2) - \frac{(1+\varepsilon)^3}{6}(a_1\varepsilon)^2 + O(\varepsilon^3) = 0$.

Collecting powers of $\varepsilon$:

- Order $\varepsilon^1$: $-1 + \frac{a_1}{2} = 0 \Rightarrow a_1 = 2$.
- Order $\varepsilon^2$: $\frac{a_2}{2} + \frac{a_1}{2}\cdot 1 - \frac{a_1^2}{6} = 0 \Rightarrow \frac{a_2}{2} + 1 - \frac{2}{3} = 0 \Rightarrow a_2 = -\frac{2}{3}$.

So $S^* \approx 2\varepsilon - \frac{2}{3}\varepsilon^2 + O(\varepsilon^3)$.

**Part 3.**

```{r}
#| label: ex3-near-threshold
#| echo: true

k_val <- 1.2
eps   <- k_val - 1
N     <- 1000
p_val <- k_val / (N - 1)

# (a) Linear approximation
S_linear <- 2 * eps
# (b) Exact numerical solution
S_exact <- solve_S(k_val)
# (c) Simulation
set.seed(42)
S_sim <- mean(sapply(1:500, function(i) {
  g <- sample_gnp(N, p_val, directed=FALSE, loops=FALSE)
  max(components(g)$csize) / N
}))

cat(sprintf("k = %.1f, eps = %.1f\n", k_val, eps))
cat(sprintf("(a) Linear approx S ≈ 2*eps:    %.4f\n", S_linear))
cat(sprintf("(b) Exact self-consistency S*:   %.4f\n", S_exact))
cat(sprintf("(c) Simulation estimate:          %.4f\n", S_sim))
```

At $\varepsilon = 0.2$, the linear approximation overestimates slightly (the $-\frac{2}{3}\varepsilon^2$ correction matters). The exact numerical solution is closer to the simulation. This demonstrates that the linear approximation is valid only very near the threshold; at $\varepsilon = 0.2$ the quadratic correction is already non-negligible.

**Part 4.** In bond percolation on a 2D square lattice, the order parameter (the probability that a site belongs to the infinite cluster) near the critical bond probability $p_c = 1/2$ scales as $P_\infty \sim (p - p_c)^\beta$ with critical exponent $\beta = 5/36 \approx 0.139$ — not linear [@bollobas2001book]. The linear scaling $S \sim \varepsilon$ is specific to the mean-field (Erdős–Rényi) setting, where each vertex has many potential neighbors and spatial correlations are absent. The 2D lattice is below the upper critical dimension ($d_c = 6$ for percolation), so fluctuations matter and the exponent differs from the mean-field value $\beta = 1$.

---

## Exercise 4. The connectedness threshold

**Part 1.** The probability that a fixed vertex $v$ is isolated is $(1-p)^{N-1}$, since each of the $N-1$ potential edges is absent with probability $(1-p)$. By linearity of expectation:

$$
\mathbb{E}[\text{isolated vertices}] = N(1-p)^{N-1} \approx N e^{-p(N-1)} = N e^{-\langle k\rangle}.
$$

Setting this equal to $1$: $N e^{-\langle k\rangle} = 1$, so $e^{-\langle k\rangle} = 1/N$, giving $\langle k\rangle = \ln N$. Equivalently, $p(N-1) = \ln N$, so $p \approx \ln N / N$. $\square$

**Part 2.**

```{r}
#| label: ex4-connectedness-threshold
#| echo: true
#| fig-cap: "Empirical connectedness probability vs theoretical curve for G(500,p)."

set.seed(42)
N_val <- 500
c_vals <- seq(-3, 5, by=0.4)
p_from_c <- function(c) (log(N_val) + c) / N_val

results <- map_dfr(c_vals, function(c) {
  p <- p_from_c(c)
  prob_conn <- mean(sapply(1:300, function(i) {
    g <- sample_gnp(N_val, p, directed=FALSE, loops=FALSE)
    as.integer(is_connected(g))
  }))
  tibble(c=c, empirical=prob_conn, theoretical=exp(-exp(-c)))
})

ggplot(results, aes(x=c)) +
  geom_point(aes(y=empirical), color=course_cols$gold, size=2.5) +
  geom_line(aes(y=theoretical), color=course_cols$teal, linewidth=1.2) +
  labs(title=expression(paste("Connectedness probability vs ", c, " in p = (ln N + c)/N")),
       subtitle=sprintf("N = %d; points = simulation, line = theory exp(-exp(-c))", N_val),
       x="c", y="P(connected)")
```

The theoretical curve $e^{-e^{-c}}$ matches the simulation closely for large $N$. Deviations reflect finite-$N$ corrections.

**Part 3.**

```{r}
#| label: ex4-ratio
#| echo: true

for (N in c(1e3, 1e6, 1e9)) {
  ratio <- log(N) / 1
  cat(sprintf("N = 10^%.0f: ratio = ln(N)/1 = %.1f\n",
              log10(N), ratio))
}
```

For $N = 10^3$: ratio $\approx 6.9$. For $N = 10^6$: ratio $\approx 13.8$. For $N = 10^9$: ratio $\approx 20.7$. As networks grow, the supercritical regime (giant component but not connected) spans an increasingly wide range of $\langle k\rangle$. Most real-world networks with $\langle k\rangle$ between $3$ and $10$ sit well within this regime for large $N$.

**Part 4.**

```{r}
#| label: ex4-isolated-vs-connected
#| echo: true
#| fig-cap: "Fraction of isolated vertices and connectedness probability as functions of expected degree for G(500,p)."

set.seed(42)
N_val <- 500
k_seq <- seq(1, 12, length.out=40)

dual_tbl <- map_dfr(k_seq, function(k) {
  p <- k / (N_val - 1)
  reps <- 200
  tibble(
    k          = k,
    iso_frac   = mean(sapply(1:reps, function(i) {
      g <- sample_gnp(N_val, p, directed=FALSE, loops=FALSE)
      sum(degree(g) == 0) / N_val
    })),
    conn_prob  = mean(sapply(1:reps, function(i) {
      g <- sample_gnp(N_val, p, directed=FALSE, loops=FALSE)
      as.integer(is_connected(g))
    }))
  )
})

dual_tbl |>
  pivot_longer(cols=c(iso_frac, conn_prob), names_to="quantity", values_to="value") |>
  mutate(quantity=recode(quantity,
    iso_frac  = "Fraction isolated",
    conn_prob = "P(connected)")) |>
  ggplot(aes(x=k, y=value, color=quantity)) +
  geom_line(linewidth=1.1) +
  geom_vline(xintercept=log(N_val), color=course_cols$slate, linetype=2) +
  scale_color_manual(values=c("Fraction isolated"=course_cols$gold,
                               "P(connected)"=course_cols$teal)) +
  labs(title="Isolated vertices and connectedness",
       subtitle=sprintf("Dashed = ln(%d) = %.2f", N_val, log(N_val)),
       x=expression(langle*k*rangle), y="Value", color=NULL)
```

The two curves cross near $\langle k\rangle = \ln N$, confirming that the disappearance of isolated vertices is the binding constraint for full connectivity.

---

## Exercise 5. Structural regimes: classification and transitions

**Part 1.**

```{r}
#| label: ex5-social-network
#| echo: true

N_soc <- 1e5; k_soc <- 3.2
S_soc <- solve_S(k_soc)
E_iso <- N_soc * exp(-k_soc)
conn_threshold <- log(N_soc)

cat(sprintf("N = %g, <k> = %.1f\n", N_soc, k_soc))
cat(sprintf("Regime: supercritical (<k> = %.1f > 1, < ln N = %.1f)\n",
            k_soc, conn_threshold))
cat(sprintf("Theoretical S* = %.4f (%.0f vertices)\n", S_soc, S_soc*N_soc))
cat(sprintf("Expected isolated vertices = %.1f\n", E_iso))
cat(sprintf("Connected? Likely NO (ln N = %.1f >> <k> = %.1f)\n",
            conn_threshold, k_soc))
```

With $\langle k\rangle = 3.2$, the network is firmly supercritical: about $96\%$ of vertices are in the giant component. However, with $\ln(10^5) \approx 11.5$ far above $3.2$, the network is not connected — roughly $N e^{-3.2} \approx 4086$ isolated vertices are expected.

**Part 2.** Under the ER model with $\langle k\rangle = 4$, the self-consistency equation gives a unique $S^* \approx 0.98$. Both Network A ($S = 0.85$) and Network B ($S = 0.43$) differ substantially from the ER prediction. Network B in particular ($S = 0.43$) is far below the ER baseline, suggesting strong clustering or community structure that fragments the network into multiple large components — the clustering reduces the effective reach of the giant component [@newman2010, Ch. 13]. Network A is also below the ER prediction, suggesting moderate non-ER structure.

**Part 3.**

```{r}
#| label: ex5-second-largest
#| echo: true
#| fig-cap: "Distribution of second-largest component size for N=1000 and three values of expected degree."

set.seed(42)
N_val2 <- 1000
k_test <- c(1.5, 2.0, 3.0)
log_n  <- log(N_val2)

second_largest_tbl <- map_dfr(k_test, function(k) {
  p <- k / (N_val2 - 1)
  map_dfr(1:100, function(i) {
    g    <- sample_gnp(N_val2, p, directed=FALSE, loops=FALSE)
    comp <- sort(components(g)$csize, decreasing=TRUE)
    tibble(k=k, second=ifelse(length(comp) >= 2, comp[2], 0))
  })
})

ggplot(second_largest_tbl, aes(x=second, fill=factor(k))) +
  geom_histogram(bins=20, alpha=0.6, position="identity") +
  geom_vline(xintercept=log_n, linetype=2, color=course_cols$slate) +
  facet_wrap(~k, labeller=label_both, scales="free_x") +
  scale_fill_manual(values=c(course_cols$teal, course_cols$gold, course_cols$coral)) +
  labs(title=sprintf("Second-largest component (N=%d)", N_val2),
       subtitle=sprintf("Dashed = ln(N) = %.1f", log_n),
       x="Second-largest component size", y="Count", fill="<k>") +
  theme(legend.position="none")
```

The second-largest component sizes concentrate near or below $\ln N \approx 6.9$, consistent with the $O(\log N)$ prediction. At $\langle k\rangle = 3$, the second-largest component is very small (almost always $< 10$), confirming that the giant component absorbs nearly all the structure.

**Part 4.**

```{r}
#| label: ex5-critical-scaling
#| echo: true
#| fig-cap: "Mean largest component size at the critical point k=1, plotted against N on a log-log scale. The slope should be 2/3."

set.seed(42)
N_crit <- c(100, 500, 1000, 5000)

crit_tbl <- map_dfr(N_crit, function(N) {
  p <- 1 / (N - 1)
  mean_max <- mean(sapply(1:200, function(i) {
    g <- sample_gnp(N, p, directed=FALSE, loops=FALSE)
    max(components(g)$csize)
  }))
  tibble(N=N, mean_max=mean_max)
})

# Fit power law
lm_fit <- lm(log(mean_max) ~ log(N), data=crit_tbl)
exponent <- coef(lm_fit)[2]
cat(sprintf("Estimated scaling exponent: %.3f (theoretical: 2/3 = 0.667)\n", exponent))

ggplot(crit_tbl, aes(x=N, y=mean_max)) +
  geom_point(color=course_cols$gold, size=3) +
  geom_smooth(method="lm", color=course_cols$teal, se=FALSE) +
  scale_x_log10() + scale_y_log10() +
  labs(title=expression(paste("Critical scaling: E[|C"[max]*"|] ~ N^(2/3) at ", langle*k*rangle, " = 1")),
       subtitle=sprintf("Estimated exponent = %.3f", exponent),
       x="N (log scale)", y="Mean largest component (log scale)")
```

The estimated exponent is close to $2/3$, consistent with the theoretical prediction [@erdos1960; @bollobas2001book]. Deviations at small $N$ reflect finite-size corrections.

---

## Exercise 6. Finite-size effects and convergence

**Part 1.**

```{r}
#| label: ex6-finite-size
#| echo: true
#| fig-cap: "Giant component fraction for different N values. The transition sharpens as N increases."

set.seed(42)
N_vals <- c(100, 500, 1000, 5000)
k_seq  <- seq(0.1, 4.0, length.out=50)

fs_tbl <- map_dfr(N_vals, function(N) {
  map_dfr(k_seq, function(k) {
    p <- k / (N - 1)
    S_mean <- mean(sapply(1:80, function(i) {
      g <- sample_gnp(N, p, directed=FALSE, loops=FALSE)
      max(components(g)$csize) / N
    }))
    tibble(N=N, k=k, S=S_mean)
  })
})

ggplot(fs_tbl, aes(x=k, y=S, color=factor(N))) +
  geom_line(linewidth=0.9) +
  geom_vline(xintercept=1, linetype=2, color=course_cols$slate) +
  scale_color_manual(values=c(course_cols$gold, course_cols$teal,
                               course_cols$blue, course_cols$coral)) +
  labs(title="Finite-size effects on the giant component transition",
       x=expression(langle*k*rangle), y="Giant component fraction S",
       color="N")
```

```{r}
#| label: ex6-sharpness
#| echo: true

# Width of transition: interval where S goes from 0.1 to 0.9
sharpness_tbl <- fs_tbl |>
  group_by(N) |>
  summarise(
    k_lo = approx(S, k, xout=0.1)$y,
    k_hi = approx(S, k, xout=0.9)$y,
    width = k_hi - k_lo,
    .groups="drop"
  )
print(sharpness_tbl)
```

The transition width decreases as $N$ increases, consistent with the sharpening predicted by theory. Qualitatively, this reflects the law of large numbers: in larger systems, fluctuations in the component structure are proportionally smaller, so the transition from fragmented to giant-component-dominated structure is more abrupt.

**Part 2.**

```{r}
#| label: ex6-empirical-threshold
#| echo: true
#| fig-cap: "Empirical threshold k_hat vs N, approaching 1 as N grows."

set.seed(42)
N_seq <- c(50, 100, 250, 500, 1000)
k_fine <- seq(0.5, 2.5, length.out=60)

threshold_tbl <- map_dfr(N_seq, function(N) {
  S_vals_loc <- sapply(k_fine, function(k) {
    p <- k / (N - 1)
    mean(sapply(1:100, function(i) {
      g <- sample_gnp(N, p, directed=FALSE, loops=FALSE)
      max(components(g)$csize) / N
    }))
  })
  k_hat <- k_fine[which.min(abs(S_vals_loc - 0.5))]
  tibble(N=N, k_hat=k_hat)
})

print(threshold_tbl)

ggplot(threshold_tbl, aes(x=N, y=k_hat)) +
  geom_point(color=course_cols$gold, size=3) +
  geom_hline(yintercept=1, linetype=2, color=course_cols$teal) +
  scale_x_log10() +
  labs(title="Empirical giant-component threshold vs N",
       subtitle="Dashed line: theoretical threshold k = 1",
       x="N (log scale)", y=expression(hat(k)[c]))
```

The empirical threshold converges toward $1$ as $N$ increases, but from above — for finite $N$, the observed transition point is slightly above $1$ because the giant component needs more than infinitesimal density to appear in a finite system.

**Part 3.** The fluctuations in $|C_{\max}|/N$ around its expectation are of order $N^{-1/3}$ near the critical point (the standard deviation of the order parameter scales as $N^{-1/3}$ at criticality). By the law of large numbers analogy: just as the sample mean of i.i.d. random variables concentrates at rate $1/\sqrt{n}$, the giant component fraction concentrates around its expected value, with concentration improving as $N$ grows. In the critical regime, the relevant scale is $N^{2/3}$ (the size of the largest critical component), so fluctuations as a fraction of $N$ are of order $N^{-1/3}$, which vanishes as $N \to \infty$.

---

## Exercise 7. The ER model as a null model for connectivity

**Part 1.**

```{r}
#| label: ex7-null-model
#| echo: true

N_obs <- 500; M_obs <- 800; S_obs <- 480/500

k_obs <- 2 * M_obs / N_obs
p_fit <- k_obs / (N_obs - 1)
S_theory <- solve_S(k_obs)

cat(sprintf("Observed: N=%d, M=%d, S_obs=%.3f\n", N_obs, M_obs, S_obs))
cat(sprintf("Fitted: <k>=%.3f, p=%.5f\n", k_obs, p_fit))
cat(sprintf("Theoretical S* = %.4f\n", S_theory))
cat(sprintf("Observed S = %.4f\n", S_obs))
cat(sprintf("Discrepancy: observed - theory = %.4f\n", S_obs - S_theory))
```

The theoretical $S^* \approx 0.97$ is close to the observed $0.96$, suggesting the connectivity structure is roughly consistent with an ER model at this density.

**Part 2.**

```{r}
#| label: ex7-simulation
#| echo: true
#| fig-cap: "Distribution of giant component fraction in 500 G(N,p) realisations. Red line: observed value."

set.seed(42)
sim_S_null <- sapply(1:500, function(i) {
  g <- sample_gnp(N_obs, p_fit, directed=FALSE, loops=FALSE)
  max(components(g)$csize) / N_obs
})

ggplot(tibble(S=sim_S_null), aes(x=S)) +
  geom_histogram(bins=30, fill=course_cols$blue, alpha=0.6, color="white") +
  geom_vline(xintercept=S_obs, color="#EF4444", linewidth=1.2, linetype=2) +
  labs(title="Giant component fraction under ER null model",
       subtitle=sprintf("Red = observed S = %.3f; mean simulation = %.3f",
                        S_obs, mean(sim_S_null)),
       x="Giant component fraction", y="Count")
```

**Part 3.**

```{r}
#| label: ex7-diagnostics
#| echo: true

E_iso        <- N_obs * exp(-k_obs)
conn_prob_th <- exp(-exp(-(k_obs - log(N_obs))))
# Second largest: O(log N) under ER
E_second     <- log(N_obs)

cat(sprintf("(a) Expected isolated vertices: %.2f\n", E_iso))
cat(sprintf("(b) P(connected) approx: %.4f\n", conn_prob_th))
cat(sprintf("(c) Expected second-largest component: O(ln N) = O(%.1f)\n", E_second))

# Simulate second-largest
second_largest_sim <- sapply(1:500, function(i) {
  g    <- sample_gnp(N_obs, p_fit, directed=FALSE, loops=FALSE)
  comp <- sort(components(g)$csize, decreasing=TRUE)
  if (length(comp) >= 2) comp[2] else 0
})
cat(sprintf("Simulated mean second-largest: %.2f\n", mean(second_largest_sim)))
```

**Part 4.** High clustering means many edges are concentrated within tight triangular neighborhoods rather than distributed globally. In a highly clustered network, the edges that form triangles are "wasted" from the perspective of long-range connectivity: they link vertices that are already close. The giant component grows by absorbing distant parts of the graph through long-range connections; when $p$ edges are concentrated locally (in triangles), fewer are available to bridge distant parts of the network [@newman2010, Ch. 13]. Equivalently, the effective number of neighbors that a vertex can reach in $r$ steps is suppressed by clustering: neighbors share connections with each other rather than exploring new vertices, reducing the branching factor of the BFS tree used to build the giant component.
