---
title: "Small-World Networks — Solutions"
subtitle: "Corresponds to `chapters/05-small-world-graphs.qmd`"
---

```{r}
#| label: solution-setup
#| include: false

setup_candidates <- c("../../R/book-setup.R", "../R/book-setup.R", "R/book-setup.R")
setup_file <- setup_candidates[file.exists(setup_candidates)][1]

if (is.na(setup_file)) {
  stop("Could not locate R/book-setup.R", call. = FALSE)
}

source(setup_file, local = knitr::knit_global())
ns_book_setup(
  preset = "smallworld",
  packages = c(ns_package_profile("core"), "tibble", "scales"),
  seed = 2051
)
```

## Exercise 1. From branching to logarithmic distance

**Part 1.** The sum is a finite geometric series with first term $1$ and ratio $\langle k \rangle \neq 1$:

$$
N(d)
=
\sum_{j=0}^{d}\langle k \rangle^{j}
=
\frac{\langle k \rangle^{d+1}-1}{\langle k \rangle-1}.
$$

The modeling assumption is a **tree-like branching approximation**: each new neighborhood layer multiplies the previous layer by $\langle k \rangle$, with no overlap between neighborhoods, no loops, and no boundary effects from the finite vertex set. Equivalently, the local exploration is treated as a branching process on an infinite tree with offspring mean $\langle k \rangle$.

**Part 2.** Set $N(d_{\max})\approx N$. For large $N$ and $\langle k \rangle > 1$, the last term dominates the geometric sum, so

$$
\langle k \rangle^{d_{\max}}\approx N.
$$

Taking natural logarithms of both sides,

$$
d_{\max}\ln\langle k \rangle \approx \ln N,
$$

and dividing by $\ln\langle k \rangle > 0$ yields

$$
d_{\max}\approx\frac{\ln N}{\ln\langle k \rangle}.
$$

The same scaling is then used as a heuristic for the average path length $L(G)$ in random-like networks [@newman2010; @strogatz2001].

**Part 3.** The estimate is not exact for three related reasons.

- **Neighborhood overlap.** In a real graph, two vertices at distance $d$ from a source may share neighbors. The branching count therefore overcounts reachable vertices once balls start to collide.
- **Loops and finite size.** Cycles return the walk to previously visited vertices, and the exploration must stop when the ball saturates the finite set $V$. Both effects cut off pure exponential growth.
- **Average versus diameter.** The derivation targets a covering scale $d_{\max}$ closer to a diameter than to the mean pairwise distance. Even when $D(G)$ scales like $\ln N / \ln\langle k \rangle$, the average $L(G)$ can differ by a slowly varying factor; the logarithmic law is a scaling statement, not an identity.

**Part 4.** A constant bound such as "exactly six steps for every pair" would mean $L(G)=O(1)$ as $N\to\infty$. Logarithmic growth $L(G)\sim c\ln N$ still grows, but extremely slowly: doubling $N$ only adds a constant $c\ln 2$. So "a few steps" in a huge system is consistent with logarithmic scaling, not with a universal fixed integer independent of size.

---

## Exercise 2. Lattice scaling and the surprise of small worlds

**Part 1.** Pack $N$ vertices into a $d$-dimensional cube (or torus) of side length $\ell$. Then $N\sim \ell^d$, so $\ell\sim N^{1/d}$. Graph distance on a regular lattice is proportional to Euclidean (or $\ell_1$) separation measured in lattice spacings. Typical and maximal separations are both of order $\ell$, hence

$$
L(G)\sim D(G)\sim N^{1/d}.
$$

**Part 2.** For $N=10^6$:

| Expression | Value |
|:---|---:|
| $\ln N$ | $\ln(10^6)\approx 13.8$ |
| $N^{1/3}$ | $100$ |
| $N^{1/2}$ | $1000$ |
| $N$ | $10^6$ |

```{r}
#| label: ex2-growth-laws
#| echo: true

N <- 1e6
tibble::tibble(
  law = c("ln N", "N^(1/3)", "N^(1/2)", "N"),
  value = c(log(N), N^(1/3), N^(1/2), N)
)
```

The logarithmic value is an order of magnitude smaller than the three-dimensional lattice scale, two orders smaller than the two-dimensional scale, and many orders smaller than the one-dimensional scale. Polynomial lattice laws and the logarithmic random-graph law are therefore qualitatively different families.

**Part 3.** Lattice intuition says that covering a large geometric system requires a path whose length tracks the linear size $N^{1/d}$. Logarithmic distances mean that the number of reachable vertices grows roughly exponentially with radius, so the same $N$ is covered in far fewer steps. The surprise of the small-world effect is not that paths are finite, but that they are dramatically shorter than low-dimensional geometry would suggest [@watts1998; @strogatz2001].

---

## Exercise 3. The regular ring lattice: degree, distance, and clustering

**Part 1.** By definition, vertex $i$ is joined to $\{i\pm 1,\ldots,i\pm K/2\}$ (indices mod $N$). These are $K$ distinct neighbors whenever $N > K$, so every degree equals $K$. The handshaking lemma then gives

$$
2|E|=\sum_v k_v = NK \implies |E|=\frac{NK}{2},
$$

and the average degree is $\langle k \rangle = 2|E|/N = K$.

**Part 2.** On the ring, the farthest vertices from a given source lie roughly opposite on the circle, at cyclic distance about $N/2$. Each step along the lattice can advance by at most $K/2$ positions (the farthest clockwise neighbor). The number of steps needed is therefore about

$$
\frac{N/2}{K/2}=\frac{N}{K}.
$$

Averaging over pairs rather than only antipodal pairs replaces $N/K$ by a constant factor of the same order, conventionally written

$$
L(0)\approx\frac{N}{2K}.
$$

This is a heuristic scaling estimate, not an exact diameter formula.

**Part 3.** Fix vertex $0$ and write $m=K/2$. Its neighborhood is

$$
\Gamma(0)=\{-m,\ldots,-1,1,\ldots,m\}.
$$

Assume $N$ is large enough that cyclic distances among these neighbors coincide with ordinary absolute differences. Two neighbors $j<k$ are adjacent iff $k-j\le m$.

- Both positive: number of pairs $=\binom{m}{2}$.
- Both negative: likewise $\binom{m}{2}$.
- Opposite signs: for each $a=1,\ldots,m$, the negative neighbor $-a$ connects to positive neighbors $k$ with $k\le m-a$, giving $\sum_{a=1}^{m-1}(m-a)=m(m-1)/2$ edges.

Total edges among neighbors:

$$
2\binom{m}{2}+\frac{m(m-1)}{2}
=
m(m-1)+\frac{m(m-1)}{2}
=
\frac{3}{2}m(m-1)
=
\frac{3K(K-2)}{8}.
$$

The local clustering coefficient is this count divided by $\binom{K}{2}$:

$$
C(0)
=
\frac{3K(K-2)/8}{K(K-1)/2}
=
\frac{3(K-2)}{4(K-1)}.
$$

Because the lattice is vertex-transitive, every vertex has the same $C_i$, so the global clustering equals this value.

**Part 4.**

```{r}
#| label: ex3-clustering-values
#| echo: true

C0 <- function(K) 3 * (K - 2) / (4 * (K - 1))
tibble::tibble(
  K = c(4, 6, 10),
  C0 = C0(c(4, 6, 10))
)
```

As $K\to\infty$, $C(0)\to 3/4$. Even a very wide local neighborhood on the ring remains far from a clique: many pairs of neighbors are still too far apart (on the ring metric) to be linked, so clustering saturates below $1$.

---

## Exercise 4. Shortcuts, clustering, and normalized observables

**Part 1.** Define

$$
\frac{C(\beta)}{C(0)},
\qquad
\frac{L(\beta)}{L(0)}.
$$

Absolute values of $C$ and $L$ depend on $N$ and $K$. Normalizing by the lattice baselines removes that dependence and makes the *shape* of the order-to-randomness transition comparable across parameter choices: one sees how quickly each quantity leaves its ordered starting point as $\beta$ grows.

**Part 2.** In the ring lattice, traveling between distant arcs requires many local hops. A single rewired long-range edge connects two previously remote regions and collapses many shortest paths that used to traverse the long arc — so $L$ drops sharply. Clustering, by contrast, is a local triangle count. If only a small fraction of edges is rewired, most neighborhoods still look like lattice neighborhoods, so most triangles survive and $C$ decreases slowly. This is the core asymmetry of the Watts--Strogatz mechanism [@watts1998].

**Part 3.** In the small-world regime one expects

$$
\frac{L(\beta)}{L(0)}\ \text{substantially smaller than }1,
\qquad
\frac{C(\beta)}{C(0)}\ \text{still close to }1,
$$

while clustering remains much larger than in a comparable random graph. That is exactly @def-small-world-regime: short global distances coexisting with retained local order.

**Part 4.** Short paths alone do not identify the model. An Erdős--Rényi graph achieves small $L$ by making edges globally random, which destroys local order and yields $C\approx p$ (typically tiny in the sparse regime). A Watts--Strogatz graph in the small-world window keeps most lattice neighborhoods intact, so clustering stays high while shortcuts restore global reachability. The models agree on one metric property and disagree on mesoscale structure.

---

## Exercise 5. Interpreting small-worldness and the limits of Watts--Strogatz

**Part 1.** Write

$$
\sigma=\frac{C/C_{\mathrm{rand}}}{L/L_{\mathrm{rand}}}.
$$

The inequality $\sigma>1$ means that, relative to a random null model with similar size and mean degree, clustering is elevated *more* than path length is inflated. Equivalently: the network is much more clustered than random, while distances remain comparable to random. Clustering alone is insufficient because a pure lattice also has very high $C$, yet fails the small-world test through large $L$ (which drives $\sigma$ down via a large $L/L_{\mathrm{rand}}$).

**Part 2.** Watts--Strogatz rewiring preserves the number of edges and only mildly perturbs degrees: most vertices stay near degree $K$. The model therefore cannot generate hubs or heavy-tailed degree distributions. Degree-dependent clustering (e.g., low-degree vertices in dense cores, hubs bridging communities) likewise requires heterogeneity or hierarchical growth mechanisms absent from homogeneous rewiring. The missing ingredients are growth and preferential attachment — the subject of the next chapter [@albert2002; @barabasi2016].

**Part 3.** A large $\sigma$ is evidence of the *phenomenology* high-$C$/low-$L$ relative to an ER null model. It does not identify a generative mechanism. Many processes (community structure, geographic networks with occasional long ties, hierarchical models, etc.) can produce $\sigma>1$. Claiming that $\sigma=4.2$ "proves" Watts--Strogatz rewiring confuses a diagnostic index with a unique causal model.

---

## Exercise 6. Distance scaling in random networks

```{r}
#| label: ex6-distance-scaling
#| echo: true

set.seed(2056)

mean_k <- 6
N_grid <- c(50, 100, 200, 400, 800)
reps <- 12

sample_connected_gnp <- function(n, p, max_iter = 200) {
  for (i in seq_len(max_iter)) {
    g <- igraph::sample_gnp(n, p, directed = FALSE, loops = FALSE)
    if (igraph::components(g)$no == 1) return(g)
  }
  stop("Could not sample a connected G(N,p).")
}

scaling_tbl <- purrr::map_dfr(N_grid, function(n) {
  p <- mean_k / (n - 1)
  vals <- replicate(reps, {
    g <- sample_connected_gnp(n, p)
    igraph::mean_distance(g, directed = FALSE, unconnected = FALSE)
  })
  tibble::tibble(
    N = n,
    L_emp = mean(vals),
    L_sd = sd(vals),
    L_theory = log(n) / log(mean_k)
  )
})

scaling_tbl
```

```{r}
#| label: ex6-distance-scaling-plot
#| echo: true
#| fig-cap: "Empirical average path length in connected ER graphs with mean degree 6, compared with ln N / ln 6."

ggplot(scaling_tbl, aes(x = N)) +
  geom_line(aes(y = L_emp, color = "Empirical"), linewidth = 1.1) +
  geom_point(aes(y = L_emp, color = "Empirical"), size = 2.2) +
  geom_errorbar(aes(ymin = L_emp - L_sd, ymax = L_emp + L_sd, color = "Empirical"),
                width = 12, linewidth = 0.7) +
  geom_line(aes(y = L_theory, color = "ln N / ln <k>"),
            linewidth = 1.1, linetype = 2) +
  scale_color_manual(values = c(
    "Empirical" = course_cols$blue,
    "ln N / ln <k>" = course_cols$gold
  )) +
  labs(
    title = "Distance scaling in random graphs",
    subtitle = sprintf("Connected G(N, p) with <k> = %d; %d replicates per N", mean_k, reps),
    x = "N",
    y = "Average path length",
    color = NULL
  )
```

**Interpretation.** The empirical curve tracks the logarithmic prediction and grows slowly with $N$. Agreement typically improves as $N$ increases, because the branching heuristic becomes more accurate in large sparse graphs and finite-size boundary effects weaken. At small $N$, neighborhood overlap sets in quickly, connectivity conditioning biases the sample toward slightly more cohesive graphs, and the prefactor hidden in "$\approx$" matters more — so modest vertical offsets are expected.

---

## Exercise 7. Identifying and quantifying the small-world regime

```{r}
#| label: ex7-ws-helpers
#| echo: true

sample_connected_ws <- function(size, K, beta, max_iter = 100) {
  stopifnot(K %% 2 == 0)
  for (i in seq_len(max_iter)) {
    g <- igraph::sample_smallworld(dim = 1, size = size, nei = K / 2, p = beta)
    if (igraph::components(g)$no == 1) return(g)
  }
  stop("Could not generate a connected Watts--Strogatz graph.")
}
```

**Part 1.** Normalized transition scan.

```{r}
#| label: ex7-ws-scan
#| echo: true

set.seed(2057)

n_ws <- 400
K_ws <- 8
beta_grid <- unique(c(0, 10^seq(-4, 0, length.out = 16)))
reps <- 6

ws_tbl <- purrr::map_dfr(beta_grid, function(beta) {
  purrr::map_dfr(seq_len(reps), function(rep_id) {
    g <- sample_connected_ws(size = n_ws, K = K_ws, beta = beta)
    tibble::tibble(
      beta = beta,
      rep = rep_id,
      clustering = igraph::transitivity(g, type = "average"),
      avg_distance = igraph::mean_distance(g, directed = FALSE, unconnected = FALSE)
    )
  })
})

ws_summary <- ws_tbl |>
  dplyr::group_by(beta) |>
  dplyr::summarise(
    clustering = mean(clustering),
    avg_distance = mean(avg_distance),
    .groups = "drop"
  )

C0 <- ws_summary$clustering[ws_summary$beta == 0]
L0 <- ws_summary$avg_distance[ws_summary$beta == 0]

ws_long <- ws_summary |>
  dplyr::mutate(
    clustering_norm = clustering / C0,
    distance_norm = avg_distance / L0
  ) |>
  tidyr::pivot_longer(
    cols = c(clustering_norm, distance_norm),
    names_to = "quantity",
    values_to = "value"
  ) |>
  dplyr::mutate(
    quantity = dplyr::recode(
      quantity,
      clustering_norm = "C(beta) / C(0)",
      distance_norm = "L(beta) / L(0)"
    )
  )
```

```{r}
#| label: ex7-ws-normalized-plot
#| echo: true
#| fig-cap: "Normalized path length drops quickly with beta while normalized clustering remains high over a wider interval — the small-world window."

ggplot(ws_long, aes(x = beta, y = value, color = quantity)) +
  geom_line(linewidth = 1.15) +
  geom_point(size = 1.8) +
  scale_x_log10() +
  scale_color_manual(values = c(
    "C(beta) / C(0)" = course_cols$teal,
    "L(beta) / L(0)" = course_cols$gold
  )) +
  labs(
    title = "Detecting the small-world regime",
    subtitle = sprintf("N = %d, K = %d; averages over %d connected replicates", n_ws, K_ws, reps),
    x = "Rewiring probability, beta",
    y = "Normalized quantity",
    color = NULL
  )
```

The small-world regime is the interval of $\beta$ in which $L(\beta)/L(0)$ has already fallen substantially while $C(\beta)/C(0)$ remains comparatively close to one. In a typical scan this appears at small but positive $\beta$ (often roughly $10^{-3}$ to a few $10^{-2}$, depending on $N$ and $K$).

**Part 2.** Three-model comparison.

```{r}
#| label: ex7-three-model-comparison
#| echo: true

set.seed(2058)

n_cmp <- 400
K_cmp <- 8
p_cmp <- K_cmp / (n_cmp - 1)
beta_sw <- 0.02

repeat {
  g_er <- igraph::sample_gnp(n_cmp, p_cmp, directed = FALSE, loops = FALSE)
  if (igraph::components(g_er)$no == 1) break
}

g_ring <- igraph::make_lattice(dimvector = n_cmp, nei = K_cmp / 2, circular = TRUE)
g_ws <- sample_connected_ws(size = n_cmp, K = K_cmp, beta = beta_sw)

network_stats <- function(g, label, g_rand) {
  C <- igraph::transitivity(g, type = "average")
  L <- igraph::mean_distance(g, directed = FALSE, unconnected = FALSE)
  C_rand <- igraph::transitivity(g_rand, type = "average")
  L_rand <- igraph::mean_distance(g_rand, directed = FALSE, unconnected = FALSE)
  tibble::tibble(
    model = label,
    clustering = C,
    avg_distance = L,
    sigma = (C / C_rand) / (L / L_rand)
  )
}

comparison_tbl <- dplyr::bind_rows(
  network_stats(g_ring, "Ring lattice", g_er),
  network_stats(g_ws, "Watts--Strogatz", g_er),
  network_stats(g_er, "Random graph", g_er)
) |>
  dplyr::mutate(
    model = factor(model, levels = c("Ring lattice", "Watts--Strogatz", "Random graph"))
  )

comparison_tbl
```

```{r}
#| label: ex7-comparison-plot
#| echo: true
#| fig-cap: "Clustering, average path length, and small-worldness index for ring lattice, Watts--Strogatz, and ER graphs of comparable size and mean degree."
#| fig-width: 10
#| fig-height: 4.2

comparison_long <- comparison_tbl |>
  tidyr::pivot_longer(
    cols = c(clustering, avg_distance, sigma),
    names_to = "measure",
    values_to = "value"
  ) |>
  dplyr::mutate(
    measure = factor(
      measure,
      levels = c("clustering", "avg_distance", "sigma"),
      labels = c("Clustering", "Average path length", "Small-worldness sigma")
    )
  )

ggplot(comparison_long, aes(x = model, y = value, fill = model)) +
  geom_col(width = 0.7, color = "white") +
  facet_wrap(~measure, scales = "free_y") +
  scale_fill_manual(values = c(
    "Ring lattice" = course_cols$teal,
    "Watts--Strogatz" = course_cols$gold,
    "Random graph" = course_cols$blue
  )) +
  labs(
    title = "Three structural regimes at comparable scale",
    x = NULL,
    y = NULL
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 15, hjust = 1)
  )
```

**Part 3.** The ring lattice has high clustering but large distances. The ER graph has short distances but weak clustering; its $\sigma$ is identically $1$ by construction when the null model is itself. The Watts--Strogatz graph in the small-$\beta$ window inherits much of the lattice clustering while its path length approaches the random-graph scale, producing $\sigma > 1$. The index $\sigma$ therefore summarizes the joint pattern (elevated $C$ relative to random, without a matching inflation of $L$) that neither $C$ nor $L$ alone identifies.
