---
title: "Graph Theory"
subtitle: "The mathematical language of network structure"
bibliography: ../references.bib
csl: ../apa.csl
---

```{r}
#| label: chapter-setup-graphtheory
#| include: false

setup_candidates <- c("R/book-setup.R", "../R/book-setup.R")
setup_file <- setup_candidates[file.exists(setup_candidates)][1]

if (is.na(setup_file)) {
  stop("Could not locate R/book-setup.R", call. = FALSE)
}

source(setup_file, local = knitr::knit_global())

ns_book_setup(
  preset = "core",
  seed = 42
)
```

## Introduction

Network science rests on a mathematical foundation: **graph theory**. Before studying random graphs, small-world structure, or scale-free degree heterogeneity, it is essential to establish the formal language used to describe networks precisely.

Graph theory is a branch of discrete mathematics concerned with collections of objects and the pairwise relations between them. Its origins trace to Leonhard Euler's 1736 solution to the *Königsberg bridge problem*, in which the question of whether a city could be traversed by crossing each of its seven bridges exactly once was reformulated as a purely combinatorial question about a drawing of vertices and edges. Euler's insight — that only the pattern of connections, not the physical geography, determines the answer — is the founding idea of graph theory [@euler1736].

In the context of this course, graph theory provides:

-   a precise vocabulary for describing network structure,
-   algebraic representations that connect combinatorics to linear algebra,
-   combinatorial results that govern global properties such as connectivity and reachability,
-   and the rigorous foundation on which the probabilistic models of later chapters are built [@diestel2017; @west2001; @newman2010; @barabasi2016].

Standard graduate references for the material in this chapter are @diestel2017 for pure graph theory and @west2001 for a combinatorial treatment. For the network science perspective that motivates the choice of topics, see @newman2010 and @barabasi2016. Spectral properties of graphs, covered in @sec-matrix-representations, are treated in depth by @chung1997.

## Learning goals

By the end of this chapter, a student should be able to:

-   represent a network as $G = (V, E)$ and report its order, size, and degree structure,
-   translate between edge lists, adjacency matrices, and `igraph` objects,
-   state and apply the handshaking lemma and its immediate consequences,
-   explain what matrix powers count, and how the Laplacian spectrum encodes connectivity,
-   distinguish paths, components, cut vertices, and bridges — the vocabulary used later for robustness and percolation,
-   recognise when a graph is bipartite and apply König's criterion,
-   compute the summary quantities in @tbl-summary-table and interpret them structurally.

## Chapter roadmap

The chapter proceeds from basic objects to structural concepts. It begins with the formal definition of a graph and the main graph types, then studies degree and matrix representations, and turns to walks, connectivity, trees, and bipartite structure. It ends with a compact computational toolkit in `R`. Each theoretical definition is paired with a concrete representation that can be explored and verified computationally.

## What is a graph?

### Formal definition

::: {#def-graph}
A **graph** $G$ is an ordered pair

$$
G = (V, E)
$$

where $V$ is a finite non-empty set called the **vertex set** (or **node set**), and $E$ is a set of unordered pairs $\{u, v\}$ with $u, v \in V$ and $u \neq v$, called the **edge set**.
:::

The number of vertices is denoted $N = |V|$ and called the **order** of the graph. The number of edges is denoted $M = |E|$ and called the **size** of the graph.

Two vertices $u$ and $v$ are **adjacent** (or **neighbors**) if $\{u, v\} \in E$. An edge $\{u, v\}$ is **incident** to each of $u$ and $v$. The **neighborhood** of $v$ is the set $N(v) = \{u \in V : \{u, v\} \in E\}$.

::: {.callout-note title="Convention"}
Throughout this chapter, all graphs are assumed to be **simple** unless stated otherwise. A simple graph contains no self-loops (edges from a vertex to itself) and no multi-edges (multiple edges between the same pair of vertices).
:::

### A motivating example {#sec-a-motivating-example}

Consider the graph with vertex set $V = \{1, 2, 3, 4, 5\}$ and edge set

$$
E = \bigl\{\{1,2\},\, \{1,3\},\, \{2,4\},\, \{3,4\},\, \{4,5\}\bigr\}.
$$

This graph has order $N = 5$ and size $M = 5$. Vertex 4 is the only neighbor of vertex 5 and the only common neighbor of vertices 2 and 3 — structural features that will become precise once degree and connectivity are defined.

```{r}
#| label: simple-graph-example
#| echo: false
#| fig-cap: "A simple graph on five vertices. Vertex labels indicate node identity."

g_example <- graph_from_edgelist(
  matrix(c(1,2, 1,3, 2,4, 3,4, 4,5), ncol = 2, byrow = TRUE),
  directed = FALSE
)

layout_ex <- matrix(
  c(-1, 1,   1, 1,   -1, -0.2,   1, -0.2,   0, -1.2),
  ncol = 2, byrow = TRUE
)

par(mar = c(1, 1, 2, 1))
plot(
  g_example,
  layout = layout_ex,
  vertex.size = 28,
  vertex.color = course_cols$blue,
  vertex.frame.color = "white",
  vertex.label.color = "white",
  vertex.label.cex = 1.1,
  vertex.label.font = 2,
  edge.color = course_cols$slate,
  edge.width = 2.2,
  main = "A simple graph G = (V, E)"
)
```

## Directed and undirected graphs

The definition above treats each edge as unordered: the relation $\{u, v\}$ is symmetric. Such a graph is **undirected**. In many real networks, relations are asymmetric — a hyperlink from page $u$ to page $v$ does not imply a link in the reverse direction; a neuron may send signals to a target without receiving signals back.

::: {#def-directed-graph}
A **directed graph** (or **digraph**) is a pair $G = (V, A)$ where $V$ is the vertex set and $A$ is a set of **ordered** pairs $(u, v)$ with $u \neq v$, called **arcs**. The arc $(u, v)$ has **tail** $u$ and **head** $v$; it is distinct from $(v, u)$.
:::

```{r}
#| label: directed-vs-undirected
#| echo: false
#| fig-cap: "Left: an undirected graph. Right: a directed graph on the same vertex set. Arcs carry orientation."

g_undir <- graph_from_edgelist(
  matrix(c(1,2, 2,3, 3,4, 4,1, 1,3), ncol = 2, byrow = TRUE),
  directed = FALSE
)

g_dir <- graph_from_edgelist(
  matrix(c(1,2, 3,2, 3,4, 1,4, 1,3), ncol = 2, byrow = TRUE),
  directed = TRUE
)

lyt <- matrix(c(0,1, 1,0, 0,-1, -1,0), ncol = 2, byrow = TRUE)

old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 2), mar = c(1, 1, 3, 1))

plot(g_undir,
     layout = lyt,
     vertex.size = 26, vertex.color = course_cols$teal,
     vertex.frame.color = "white", vertex.label.color = "white",
     vertex.label.font = 2,
     edge.color = course_cols$slate, edge.width = 2,
     main = "Undirected graph")

plot(g_dir,
     layout = lyt,
     vertex.size = 26, vertex.color = course_cols$gold,
     vertex.frame.color = "white", vertex.label.color = "white",
     vertex.label.font = 2,
     edge.color = course_cols$warm, edge.width = 2,
     edge.arrow.size = 0.7,
     main = "Directed graph")

par(old_par)
```

### Weighted graphs

::: {#def-weighted-graph}
A **weighted graph** is a graph $G = (V, E)$ together with a weight function $w : E \to \mathbb{R}$, assigning a real-valued weight to each edge. Weights may represent distances, capacities, interaction strengths, or costs, depending on the application.
:::

## Degree and the handshaking lemma

### Degree of a vertex

::: {#def-degree}
Let $G = (V, E)$ be an undirected graph. The **degree** of vertex $v \in V$, denoted $k_v$ or $\deg(v)$, is the number of edges incident to $v$:

$$
k_v = |\{u \in V : \{u,v\} \in E\}| = |N(v)|.
$$
:::

A vertex with $k_v = 0$ is **isolated**; one with $k_v = 1$ is a **leaf** or **pendant vertex**. The maximum and minimum degrees of $G$ are denoted $\Delta(G)$ and $\delta(G)$ respectively.

### Degree sequence

::: {#def-degree-sequence}
The **degree sequence** of $G$ is the list of vertex degrees arranged in non-increasing order:

$$
k_1 \geq k_2 \geq \cdots \geq k_N.
$$
:::

The degree sequence is a fundamental invariant of the graph. Not every integer sequence is graphical — the Erdős–Gallai theorem gives a necessary and sufficient condition for a sequence of non-negative integers to be realizable as a degree sequence of a simple graph.

### The handshaking lemma

::: {#thm-handshaking}
**Handshaking Lemma.** For any undirected graph $G = (V, E)$,

$$
\sum_{v \in V} k_v = 2M.
$$
:::

**Proof.** Consider the incidence sum $\sum_{v \in V} \sum_{e \in E} \mathbf{1}[v \in e]$. Evaluated by vertex, it equals $\sum_{v} k_v$. Evaluated by edge, each edge $\{u,v\}$ contributes exactly $1$ to $v = u$ and exactly $1$ to $v$, so the sum equals $2M$. $\square$

::: {.callout-note title="Immediate corollary"}
Every graph has an even number of odd-degree vertices. Since $2M$ is even, the sum $\sum_v k_v$ is even; an even sum requires an even count of odd summands.
:::

### Average degree

::: {#def-average-degree}
The **average degree** of $G$ is

$$
\langle k \rangle = \frac{1}{N} \sum_{v \in V} k_v = \frac{2M}{N}.
$$
:::

The identity $\langle k \rangle = 2M/N$ follows immediately from the handshaking lemma. It connects average degree, edge count, and graph order in a single formula that recurs throughout this book.

### In-degree and out-degree in directed graphs

In a directed graph, each arc contributes asymmetrically to two vertices.

::: {#def-indegree-outdegree}
Let $G = (V, A)$ be a directed graph and $v \in V$. The **in-degree** of $v$ is

$$
k_v^{\text{in}} = |\{u \in V : (u, v) \in A\}|
$$

and the **out-degree** of $v$ is

$$
k_v^{\text{out}} = |\{u \in V : (v, u) \in A\}|.
$$
:::

The directed analogue of the handshaking lemma states:

$$
\sum_{v \in V} k_v^{\text{in}} = \sum_{v \in V} k_v^{\text{out}} = |A|.
$$

Every arc $(u, v)$ contributes $1$ to the out-degree of $u$ and $1$ to the in-degree of $v$.

```{r}
#| label: degree-distribution-example
#| echo: false
#| fig-cap: "Degree sequence (sorted in decreasing order) of a graph on 30 vertices. The dashed line marks the average degree."

set.seed(42)
g_deg <- sample_gnp(n = 30, p = 0.18)
deg_vec <- degree(g_deg)
avg_deg <- mean(deg_vec)

deg_df <- data.frame(
  vertex = seq_along(deg_vec),
  degree = sort(deg_vec, decreasing = TRUE)
)

ggplot(deg_df, aes(x = vertex, y = degree)) +
  geom_col(fill = course_cols$blue, alpha = 0.8, width = 0.8) +
  geom_hline(
    yintercept = avg_deg,
    linetype = "dashed",
    color = course_cols$warm,
    linewidth = 0.9
  ) +
  annotate(
    "text",
    x = 27,
    y = avg_deg + 0.55,
    label = paste0("'\u27e8' * k * '\u27e9' == ", round(avg_deg, 2)),
    parse = TRUE,
    color = course_cols$warm,
    size = 3.8,
    hjust = 1
  ) +
  labs(
    title = "Degree sequence (sorted)",
    x = "Vertex rank",
    y = "Degree"
  ) +
  scale_y_continuous(breaks = 0:10)
```

## Matrix representations {#sec-matrix-representations}

Matrix representations are central to network science because they translate combinatorial graph structure into linear-algebraic objects, making it possible to study connectivity, diffusion, centrality, and spectral properties computationally [@newman2010; @chung1997]. The adjacency matrix and the Laplacian are the two most important representations; a comprehensive treatment of their spectral theory is given in @chung1997.

### The adjacency matrix

::: {#def-adjacency-matrix}
Let $G = (V, E)$ be a graph with $V = \{1, 2, \ldots, N\}$. The **adjacency matrix** of $G$ is the $N \times N$ matrix $A$ defined entry-wise by

$$
A_{ij} = \begin{cases} 1 & \text{if } \{i, j\} \in E, \\ 0 & \text{otherwise.} \end{cases}
$$
:::

For an undirected graph, $A$ is **symmetric**: $A_{ij} = A_{ji}$ for all $i, j$. For a simple graph, $A_{ii} = 0$ for all $i$.

#### Key properties

Several properties follow immediately from the definition:

-   The row sum of row $i$ equals the degree of vertex $i$: $\sum_j A_{ij} = k_i$.
-   The total sum of all entries equals $2M$: $\sum_{i,j} A_{ij} = 2M$.
-   The entry $(A^r)_{ij}$ of the $r$-th matrix power counts the number of **walks** of length $r$ from vertex $i$ to vertex $j$.

The third property follows from the definition of matrix multiplication: $(A^2)_{ij} = \sum_k A_{ik} A_{kj}$ counts the number of common neighbors of $i$ and $j$, which is exactly the number of walks of length $2$ connecting them. The argument extends by induction to any $r \geq 1$.

### The degree matrix

::: {#def-degree-matrix}
The **degree matrix** $D$ is the $N \times N$ diagonal matrix with $D_{ii} = k_i$ and $D_{ij} = 0$ for $i \neq j$.
:::

### The Laplacian matrix

::: {#def-laplacian}
The **Laplacian matrix** of $G$ is

$$
L = D - A.
$$
:::

The Laplacian is symmetric and positive semi-definite. Its entries are:

$$
L_{ij} = \begin{cases} k_i & \text{if } i = j, \\ -1 & \text{if } \{i, j\} \in E, \\ 0 & \text{otherwise.} \end{cases}
$$

Two structural facts about $L$ are fundamental:

1. The smallest eigenvalue of $L$ is always $0$, with eigenvector $\mathbf{1}$ (the all-ones vector), because $L\mathbf{1} = D\mathbf{1} - A\mathbf{1} = \mathbf{k} - \mathbf{k} = \mathbf{0}$. The multiplicity of the zero eigenvalue equals the number of connected components.
2. The **Fiedler value** $\lambda_2(L)$ — the second-smallest eigenvalue, introduced by @fiedler1973 — measures how well-connected the graph is. A larger $\lambda_2$ implies greater robustness: more edge-disjoint paths exist between typical pairs of vertices [@chung1997].

```{r}
#| label: adjacency-matrix-visual
#| echo: false
#| fig-cap: "Left: a small graph on six vertices. Right: its adjacency matrix as a heatmap. Coloured cells indicate edges."

set.seed(7)
g_mat <- graph_from_edgelist(
  matrix(c(1,2, 1,3, 2,4, 3,4, 3,5, 4,6, 5,6), ncol = 2, byrow = TRUE),
  directed = FALSE
)

A <- as.matrix(as_adjacency_matrix(g_mat))
lyt_mat <- layout_with_kk(g_mat)

old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 2), mar = c(2, 2, 3, 1))

plot(g_mat,
     layout = lyt_mat,
     vertex.size = 24,
     vertex.color = course_cols$teal,
     vertex.frame.color = "white",
     vertex.label.color = "white",
     vertex.label.font = 2,
     edge.color = course_cols$slate,
     edge.width = 2,
     main = "Graph G")

N_mat <- nrow(A)
image(
  1:N_mat, 1:N_mat, A,
  col = c("white", course_cols$teal),
  xlab = "", ylab = "",
  main = "Adjacency matrix A",
  axes = FALSE
)
axis(1, at = 1:N_mat, labels = 1:N_mat, col.axis = course_cols$ink, cex.axis = 1)
axis(2, at = 1:N_mat, labels = 1:N_mat, col.axis = course_cols$ink, cex.axis = 1, las = 1)
box(col = course_cols$grid)

par(old_par)
```

## Walks, paths, and cycles

### Walk, trail, and path

::: {#def-walk}
A **walk** of length $\ell$ in $G$ is a finite sequence of vertices

$$
v_0, v_1, \ldots, v_\ell
$$

such that $\{v_{i-1}, v_i\} \in E$ for every $i = 1, \ldots, \ell$. A walk may repeat vertices and edges.
:::

::: {#def-trail-path}
A **trail** is a walk in which no edge is repeated. A **path** is a walk in which no vertex is repeated (which implies no edge is repeated).
:::

The **distance** $d(u, v)$ between two vertices is the length of a shortest path connecting them (or $\infty$ if no path exists). The **diameter** $D(G) = \max_{u,v} d(u,v)$ is the maximum distance over all pairs of vertices.

### Cycle

::: {#def-cycle}
A **cycle** is a closed walk $v_0, v_1, \ldots, v_\ell = v_0$ of length $\ell \geq 3$ in which all vertices $v_0, \ldots, v_{\ell-1}$ are distinct. The **girth** of $G$ is the length of its shortest cycle. A graph with no cycles is **acyclic**.
:::

::: {.callout-note title="Matrix powers and triangle counting"}
Because $(A^r)_{ij}$ counts walks of length $r$ from $i$ to $j$, the diagonal entry $(A^3)_{ii}$ counts closed walks of length $3$ through $i$ — that is, the number of triangles containing $i$. Summing over all vertices and dividing by $6$ (two traversal directions $\times$ three starting vertices per triangle) gives:

$$
\text{triangles}(G) = \frac{1}{6} \operatorname{tr}(A^3).
$$

This identity connects combinatorial structure to a straightforward matrix computation.
:::

## Connectivity

### Connected components

::: {#def-connectivity}
An undirected graph $G$ is **connected** if for every pair of distinct vertices $u, v \in V$, there exists a path from $u$ to $v$. A **connected component** is a maximal connected subgraph of $G$.
:::

The number of connected components is denoted $c(G)$; $c(G) = 1$ if and only if $G$ is connected. The multiplicity of the zero eigenvalue of $L$ equals $c(G)$ — providing an algebraic route to detecting components.

### Strong and weak connectivity in directed graphs

::: {#def-strong-weak-connectivity}
A directed graph is **strongly connected** if for every ordered pair $(u, v)$ there is a directed path from $u$ to $v$. It is **weakly connected** if the underlying undirected graph (obtained by ignoring arc directions) is connected.
:::

Strong connectivity requires reachability in both directions between every pair of vertices. A directed graph can be weakly connected without being strongly connected.

```{r}
#| label: connected-components-plot
#| echo: false
#| fig-cap: "A disconnected graph with three connected components shown in distinct colours. Component membership is determined by mutual reachability."

g_comp <- make_empty_graph(n = 12, directed = FALSE)
g_comp <- add_edges(g_comp, c(1,2, 2,3, 3,1,
                               4,5, 5,6, 6,7, 7,4,
                               8,9, 9,10, 10,11, 11,12, 12,8))

comp_membership <- components(g_comp)$membership
comp_cols <- c(course_cols$blue, course_cols$teal, course_cols$gold)
vertex_cols <- comp_cols[comp_membership]

par(mar = c(1, 1, 2, 1))
set.seed(10)
plot(
  g_comp,
  layout = layout_with_fr(g_comp),
  vertex.size = 22,
  vertex.color = vertex_cols,
  vertex.frame.color = "white",
  vertex.label.color = "white",
  vertex.label.font = 2,
  vertex.label.cex = 0.9,
  edge.color = course_cols$slate,
  edge.width = 2,
  main = "Three connected components"
)
legend("bottomright",
       legend = c("Component 1 (triangle)",
                  "Component 2 (4-cycle)",
                  "Component 3 (5-cycle)"),
       fill = comp_cols, border = NA,
       bty = "n", cex = 0.88, text.col = course_cols$ink)
```

### Cut vertices and bridges

Two structural elements are critical for understanding how connectivity can fail.

::: {#def-cut-vertex}
A **cut vertex** (or **articulation point**) is a vertex $v$ whose removal strictly increases the number of connected components: $c(G - v) > c(G)$.
:::

::: {#def-bridge}
A **bridge** is an edge $e$ whose removal strictly increases the number of connected components: $c(G - e) > c(G)$.
:::

Cut vertices and bridges are the most fragile elements of a connected graph. Their removal disconnects the network — a property directly relevant to network robustness, examined in detail in Chapter 7. Note that every bridge $\{u,v\}$ is contained in no cycle: if it were, its removal would leave $u$ and $v$ connected via the cycle, contradicting the definition.

```{r}
#| label: cut-vertex-bridge
#| echo: false
#| fig-cap: "A graph with a cut vertex (red) and a bridge (dashed red edge). Removing either one disconnects the graph."

g_cut <- graph_from_edgelist(
  matrix(c(1,2, 2,3, 3,1,
           3,4,
           4,5, 5,6, 6,4),
         ncol = 2, byrow = TRUE),
  directed = FALSE
)

lyt_cut <- matrix(c(-1.2,0, -0.4,0.9, -0.4,-0.9,
                     0.4, 0,
                     1.2, 0.9, 1.2,-0.9), ncol = 2, byrow = TRUE)

cut_v <- articulation_points(g_cut)
vertex_col <- rep(course_cols$teal, vcount(g_cut))
vertex_col[cut_v] <- "#EF4444"

bridge_idx <- which(apply(as_edgelist(g_cut), 1, function(e) {
  g_tmp <- delete_edges(g_cut, get.edge.ids(g_cut, e))
  !is_connected(g_tmp)
}))
edge_col <- rep(course_cols$slate, ecount(g_cut))
edge_lty <- rep(1, ecount(g_cut))
edge_col[bridge_idx] <- "#EF4444"
edge_lty[bridge_idx] <- 2

par(mar = c(1, 1, 2.5, 1))
plot(g_cut,
     layout = lyt_cut,
     vertex.size = 26,
     vertex.color = vertex_col,
     vertex.frame.color = "white",
     vertex.label.color = "white",
     vertex.label.font = 2,
     edge.color = edge_col,
     edge.lty = edge_lty,
     edge.width = 2.5,
     main = "Cut vertex (red node) and bridge (dashed red edge)")
```

## Trees and spanning trees

### Definition and characterisation

::: {#def-tree}
A **tree** is a connected acyclic graph. A **forest** is an acyclic graph (whose connected components are trees).
:::

::: {#thm-tree-equivalences}
**Theorem.** Let $G$ be a graph with $N$ vertices. The following are equivalent:

1.  $G$ is a tree.
2.  $G$ is connected and has exactly $N - 1$ edges.
3.  $G$ is acyclic and has exactly $N - 1$ edges.
4.  Any two distinct vertices are connected by exactly one path.
5.  $G$ is connected, and removing any single edge disconnects it (equivalently, every edge is a bridge).
:::

The equivalence of these statements is classical; proofs appear in @diestel2017 [Ch. 1] and @west2001 [Ch. 2]. In particular, $M = N - 1$ is both necessary and sufficient for a connected graph to be a tree. This identity will resurface in Chapter 4 when characterising the giant component of a random graph near the phase transition.

### Spanning trees

::: {#def-spanning-tree}
A **spanning tree** of a connected graph $G = (V, E)$ is a subgraph $T = (V, E')$, with $E' \subseteq E$, that is itself a tree — it contains all $N$ vertices with exactly $N-1$ edges, is connected, and is acyclic.
:::

Every connected graph possesses at least one spanning tree. When edge weights are present, the **minimum spanning tree** minimises total edge weight; the Kruskal and Prim algorithms find it in $O(M \log N)$ time.

```{r}
#| label: spanning-tree-plot
#| echo: false
#| fig-cap: "Left: a connected graph on nine vertices. Right: one spanning tree — gold edges are in the tree, grey edges are excluded."

set.seed(99)
g_span <- sample_gnp(n = 9, p = 0.42)
while (!is_connected(g_span)) {
  set.seed(sample(1:1000, 1))
  g_span <- sample_gnp(n = 9, p = 0.42)
}

spanning_tree <- mst(g_span)
sp_edges <- as_edgelist(spanning_tree, names = FALSE)
all_edges <- as_edgelist(g_span, names = FALSE)

is_tree_edge <- apply(all_edges, 1, function(e) {
  any(apply(sp_edges, 1, function(se) setequal(e, se)))
})

lyt_sp <- layout_with_kk(g_span)

old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 2), mar = c(1, 1, 3, 1))

plot(g_span,
     layout = lyt_sp,
     vertex.size = 22, vertex.color = course_cols$blue,
     vertex.frame.color = "white", vertex.label.color = "white",
     vertex.label.font = 2,
     edge.color = course_cols$slate, edge.width = 1.6,
     main = "Connected graph G")

edge_col <- ifelse(is_tree_edge, course_cols$gold, scales::alpha(course_cols$slate, 0.25))
edge_wid <- ifelse(is_tree_edge, 3.0, 1.0)

plot(g_span,
     layout = lyt_sp,
     vertex.size = 22, vertex.color = course_cols$blue,
     vertex.frame.color = "white", vertex.label.color = "white",
     vertex.label.font = 2,
     edge.color = edge_col, edge.width = edge_wid,
     main = "Spanning tree (gold edges)")

par(old_par)
```

## Bipartite graphs

### Definition and König's criterion

::: {#def-bipartite}
A graph $G = (V, E)$ is **bipartite** if $V$ can be partitioned into two non-empty disjoint sets $X$ and $Y$ such that every edge has one endpoint in $X$ and one in $Y$. The pair $(X, Y)$ is the **bipartition**.
:::

Bipartite graphs admit no edges within $X$ or within $Y$. They arise naturally in two-mode network models: authors and papers, genes and diseases, employees and projects.

::: {#thm-koenig}
**Theorem [@koenig1916].** A graph $G$ is bipartite if and only if it contains no cycle of odd length.
:::

This criterion is both elegant and computationally efficient: bipartiteness is detected by a two-coloring of $G$, executable in $O(N + M)$ time by breadth-first search [@west2001]. When BFS assigns the same color to both endpoints of any edge, an odd cycle has been found and the graph is not bipartite. Bipartite graphs arise naturally in two-mode network models — authors and papers, genes and diseases, employees and projects — and require adapted statistical tools; see @newman2010 [Ch. 6] for a discussion of projection methods.

### Complete bipartite graph

::: {#def-complete-bipartite}
The **complete bipartite graph** $K_{m,n}$ is the bipartite graph with $|X| = m$ and $|Y| = n$ in which every vertex of $X$ is adjacent to every vertex of $Y$. It has $mn$ edges, $n$ vertices of degree $m$, and $m$ vertices of degree $n$.
:::

```{r}
#| label: bipartite-plot
#| echo: false
#| fig-cap: "Left: a bipartite graph with bipartition shown in blue and teal. Right: the complete bipartite graph K(3,4)."

g_bip <- graph_from_biadjacency_matrix(
  matrix(c(1,0,1,1, 0,1,1,0, 1,1,0,1), nrow = 3, byrow = TRUE)
)

g_k34 <- make_full_bipartite_graph(3, 4)

lyt_bip <- layout_as_bipartite(g_bip)
lyt_k34 <- layout_as_bipartite(g_k34)

bip_type <- V(g_bip)$type
k34_type <- V(g_k34)$type

old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 2), mar = c(1, 1, 3, 1))

plot(g_bip,
     layout = lyt_bip[, c(2, 1)],
     vertex.size = 20,
     vertex.color = ifelse(bip_type, course_cols$teal, course_cols$blue),
     vertex.frame.color = "white",
     vertex.label.color = "white",
     vertex.label.font = 2,
     edge.color = course_cols$slate, edge.width = 2,
     main = "Bipartite graph")

plot(g_k34,
     layout = lyt_k34[, c(2, 1)],
     vertex.size = 20,
     vertex.color = ifelse(k34_type, course_cols$teal, course_cols$blue),
     vertex.frame.color = "white",
     vertex.label.color = "white",
     vertex.label.font = 2,
     edge.color = course_cols$slate, edge.width = 2,
     main = expression(K[3*","*4]))

par(old_par)
```

## Special graph families

Several families appear repeatedly in network science as benchmarks and building blocks.

::: {#def-complete-graph}
The **complete graph** $K_N$ is the simple graph on $N$ vertices in which every pair of distinct vertices is adjacent. It has $M = \binom{N}{2}$ edges and every vertex has degree $N - 1$.
:::

::: {#def-cycle-graph}
The **cycle graph** $C_N$ is a single cycle through all $N$ vertices: every vertex has degree $2$ and the graph has $N$ edges.
:::

::: {#def-path-graph}
The **path graph** $P_N$ has $N$ vertices forming a single path: the two endpoints have degree $1$, all interior vertices have degree $2$, and the graph has $N - 1$ edges.
:::

::: {#def-regular}
A graph is **$k$-regular** if every vertex has degree exactly $k$. The complete graph $K_N$ is $(N-1)$-regular; $C_N$ is $2$-regular.
:::

```{r}
#| label: special-graphs-plot
#| echo: false
#| fig-cap: "Four standard graph families: K5 (complete), C6 (cycle), a 3-regular graph on 10 vertices, and P6 (path)."

g_k5  <- make_full_graph(5)
g_c6  <- make_ring(6)
g_cub <- sample_k_regular(10, 3)
g_p6  <- make_lattice(c(6), circular = FALSE)

old_par <- par(no.readonly = TRUE)
par(mfrow = c(2, 2), mar = c(1, 1, 3, 1))

plot(g_k5, layout = layout_in_circle(g_k5),
     vertex.size = 22, vertex.color = course_cols$blue,
     vertex.frame.color = "white", vertex.label = NA,
     edge.color = course_cols$slate, edge.width = 1.8,
     main = expression(K[5] ~ "(complete)"))

plot(g_c6, layout = layout_in_circle(g_c6),
     vertex.size = 22, vertex.color = course_cols$teal,
     vertex.frame.color = "white", vertex.label = NA,
     edge.color = course_cols$slate, edge.width = 1.8,
     main = expression(C[6] ~ "(cycle)"))

plot(g_cub, layout = layout_with_fr(g_cub),
     vertex.size = 18, vertex.color = course_cols$gold,
     vertex.frame.color = "white", vertex.label = NA,
     edge.color = course_cols$slate, edge.width = 1.8,
     main = "3-regular (cubic) graph")

plot(g_p6, layout = matrix(c(1:6, rep(0,6)), ncol=2),
     vertex.size = 22, vertex.color = course_cols$warm,
     vertex.frame.color = "white", vertex.label = NA,
     edge.color = course_cols$slate, edge.width = 1.8,
     main = expression(P[6] ~ "(path)"))

par(old_par)
```

## Subgraphs and graph isomorphism

::: {#def-subgraph}
A graph $H = (V', E')$ is a **subgraph** of $G = (V, E)$ if $V' \subseteq V$ and $E' \subseteq E$. It is an **induced subgraph** on $V' \subseteq V$, written $G[V']$, if $E'$ contains every edge of $G$ with both endpoints in $V'$.
:::

::: {#def-isomorphism}
Two graphs $G = (V_1, E_1)$ and $H = (V_2, E_2)$ are **isomorphic**, written $G \cong H$, if there exists a bijection $\phi : V_1 \to V_2$ preserving adjacency:

$$
\{u, v\} \in E_1 \iff \{\phi(u), \phi(v)\} \in E_2.
$$

The function $\phi$ is a **graph isomorphism**.
:::

Isomorphic graphs are structurally identical: one is a relabeling of the other. A necessary condition for isomorphism is that the two graphs share the same degree sequence — but this is not sufficient. Constructing non-isomorphic graphs with the same degree sequence is a useful exercise in understanding what degree alone can and cannot determine.

## Working with graphs in R {#sec-working-with-graphs-in-r}

The `igraph` package provides a comprehensive toolkit for constructing, manipulating, and analysing graphs in R.

### Constructing graphs

```{r}
#| label: igraph-construction
#| echo: true

library(igraph)

# From an edge list
g1 <- graph_from_edgelist(
  matrix(c(1,2, 1,3, 2,3, 3,4, 4,5), ncol = 2, byrow = TRUE),
  directed = FALSE
)

# From an adjacency matrix
A_mat <- matrix(c(0,1,1,0,
                  1,0,1,1,
                  1,1,0,0,
                  0,1,0,0), nrow = 4)
g2 <- graph_from_adjacency_matrix(A_mat, mode = "undirected")

# Special constructors
g_cycle  <- make_ring(8)
g_full   <- make_full_graph(6)
g_random <- sample_gnp(n = 20, p = 0.20, directed = FALSE)
```

### Basic graph attributes

```{r}
#| label: igraph-attributes
#| echo: true

cat("Order (vertices):     ", vcount(g1), "\n")
cat("Size (edges):         ", ecount(g1), "\n")
cat("Degree sequence:      ",
    paste(sort(degree(g1), decreasing = TRUE), collapse = " "), "\n")
cat("Average degree:       ", round(mean(degree(g1)), 3), "\n")
cat("Is connected?:        ", is_connected(g1), "\n")
cat("Number of components: ", components(g1)$no, "\n")
cat("Diameter:             ", diameter(g1), "\n")
cat("Girth:                ", girth(g1)$girth, "\n")
```

### Adjacency and Laplacian matrices

```{r}
#| label: igraph-matrices
#| echo: true

A_out  <- as.matrix(as_adjacency_matrix(g1))
D_out  <- diag(rowSums(A_out))
L_out  <- D_out - A_out
eigs   <- round(sort(eigen(L_out)$values), 6)

cat("Adjacency matrix:\n")
print(A_out)

cat("\nLaplacian eigenvalues:\n")
print(eigs)

cat("\nFiedler value (algebraic connectivity):", eigs[2], "\n")
```

Since `g1` is connected, $L$ has exactly one zero eigenvalue. The Fiedler value $\lambda_2 > 0$ confirms connectivity; its magnitude quantifies how robustly the graph is connected.

### Triangle counting via matrix trace

```{r}
#| label: triangle-count
#| echo: true

A3    <- A_out %*% A_out %*% A_out
n_tri <- sum(diag(A3)) / 6
cat("Number of triangles (via tr(A^3)/6):", n_tri, "\n")
cat("Number of triangles (igraph)        :", sum(count_triangles(g1)) / 3, "\n")
```

## Summary of key definitions

The table below collects the most important quantities introduced in this chapter.

```{r}
#| label: tbl-summary-table
#| echo: false
#| tbl-cap: "Core graph-theoretic quantities, their notation, and formulas."

summary_df <- data.frame(
  Symbol   = c("$N = \\lvert V \\rvert$", "$M = \\lvert E \\rvert$", "$k_v$",
               "$\\langle k \\rangle$", "$A$", "$L$",
               "$d(u,v)$", "$D(G)$", "$c(G)$", "$\\lambda_2(L)$"),
  Quantity = c("Order", "Size", "Degree of $v$",
               "Average degree", "Adjacency matrix", "Laplacian matrix",
               "Distance", "Diameter", "Component count", "Fiedler value"),
  Formula  = c("$\\lvert V \\rvert$", "$\\lvert E \\rvert$",
               "$\\lvert \\{u: uv\\in E\\} \\rvert$",
               "$2M/N$", "$A_{ij}=\\mathbf{1}[ij\\in E]$", "$D-A$",
               "Length of shortest $u$-$v$ path",
               "$\\max_{u,v} d(u,v)$",
               "Maximal connected subgraph count",
               "2nd smallest eigenvalue of $L$")
)

fmt <- if (knitr::is_latex_output()) "latex" else "pipe"

knitr::kable(
  summary_df,
  format = fmt,
  align = c("c","l","l"),
  escape = FALSE,
  booktabs = knitr::is_latex_output()
)
```

## Exercises

The exercises below are designed for MSc-level students. They do not ask you to reproduce derivations from the text. Instead, they ask you to **apply, extend, and connect** the ideas introduced in this chapter. Several problems have no unique answer — the goal is precise reasoning, not a single correct number.

::: {.callout-important title="Orientation"}
These problems train you to work with the quantities in @tbl-summary-table as diagnostic tools. The same tools reappear throughout the course — in random graph theory, small-world models, and network robustness. The habit of computing, interpreting, and questioning each quantity is what this chapter is building.
:::

---

#### Exercise 1. Degree sequences: existence and uniqueness

1.  Give an example of two non-isomorphic graphs on six vertices that share the same degree sequence. Prove they are not isomorphic.
2.  State (without proof) the Erdős–Gallai condition [@erdosgallai1960] for a sequence of non-negative integers to be graphical. Apply it to determine which of the following sequences are degree sequences of a simple graph:
    (a) $(4, 3, 3, 2, 2)$,
    (b) $(5, 3, 3, 2, 1)$,
    (c) $(4, 4, 3, 3, 2, 2)$.
3.  Prove that there is no simple graph whose degree sequence is $(5, 4, 3, 2, 1)$ on five vertices. What goes wrong?

---

#### Exercise 2. The Laplacian as a diagnostic tool

Let $G$ be the graph on $\{1,2,3,4\}$ with edge set $\{\{1,2\},\{1,3\},\{2,3\},\{3,4\}\}$.

1.  Write $A$ and $L = D - A$ explicitly.
2.  Compute the eigenvalues of $L$ (numerically is fine). Verify that the smallest is $0$ and that its multiplicity matches $c(G)$.
3.  Now delete the edge $\{3,4\}$ and recompute the Laplacian spectrum. What changes, and why? Interpret the Fiedler value before and after deletion.
4.  A student claims: *"The Fiedler value tells us how many edges we need to remove to disconnect the graph."* Is this correct? Give a precise statement of what $\lambda_2$ does and does not tell us.

---

#### Exercise 3. Matrix powers and walk counting

Let $A$ be the adjacency matrix of the motivating graph from @sec-a-motivating-example.

1.  Compute $A^2$ and $A^3$ (by hand or in R).
2.  Use $(A^2)_{ij}$ to identify all pairs of vertices at distance exactly $2$. Verify against the graph directly.
3.  Use $\operatorname{tr}(A^3)/6$ to count triangles. Explain why the factor $6$ appears.
4.  The entry $(A^r)_{ij}$ counts walks, not paths. Construct a specific example (using this or any other graph) where $(A^2)_{ij} > 0$ but the number of paths of length $2$ from $i$ to $j$ is strictly less than $(A^2)_{ij}$. Explain the discrepancy.

---

#### Exercise 4. Connectivity and fragility

Consider the graph formed by two triangles, $\{1,2,3\}$ and $\{4,5,6\}$, joined by the single edge $\{3,4\}$.

1.  Identify all cut vertices and bridges. Justify each answer using the definitions in @def-cut-vertex and @def-bridge.
2.  Prove that every bridge is contained in no cycle. Use this to explain why $P_N$ has $N-2$ cut vertices and $N-1$ bridges, while $C_N$ has neither.
3.  Compute the Fiedler value for the two-triangle graph and for $C_6$ (which has the same number of vertices and edges). Which is larger? Does the Fiedler value correctly predict which graph is more fragile? Discuss any limitations of this comparison.
4.  In `igraph`, use `articulation_points()` and an edge-deletion loop to verify your classification. Report your code and output.

---

#### Exercise 5. Trees: characterisation and counting

1.  Prove that a connected graph $G$ with $N$ vertices is a tree if and only if it has exactly $N-1$ edges. Your proof should use the equivalences in @thm-tree-equivalences.
2.  How many distinct labeled trees exist on $N = 4$ vertices? List them explicitly and verify using Cayley's formula $N^{N-2}$ [@cayley1889].
3.  A graph $G$ has $N = 10$ vertices and $M = 9$ edges. A student concludes it must be a tree. Is this necessarily true? If not, give a counterexample and state the missing condition.
4.  In `igraph`, generate a random tree on 15 vertices using `sample_tree(15)`. Verify that it satisfies all five conditions in @thm-tree-equivalences computationally.

---

#### Exercise 6. Bipartite structure: theory and detection

1.  Prove König's criterion: a graph is bipartite if and only if it contains no odd cycle. *(Hint: for the forward direction, show that a 2-colouring assigns opposite colours to endpoints of every edge; for the reverse, show BFS produces a valid 2-colouring when no odd cycle exists.)*
2.  The complete bipartite graph $K_{m,n}$ has $mn$ edges. Derive a formula for its average degree $\langle k \rangle$ in terms of $m$ and $n$. For what value of $m$ (with $n$ fixed) is $\langle k \rangle$ maximised, and does this make sense intuitively?
3.  Consider the following adjacency matrix:

$$
A = \begin{pmatrix}
0 & 1 & 0 & 1 \\
1 & 0 & 1 & 0 \\
0 & 1 & 0 & 1 \\
1 & 0 & 1 & 0
\end{pmatrix}.
$$

Without drawing the graph, determine whether it is bipartite using only the eigenvalues of $A$. *(Hint: a graph is bipartite if and only if its adjacency spectrum is symmetric about zero [@chung1997].)*

4.  Give a concrete real-world example of a network that is naturally bipartite. Explain what would be structurally wrong with computing the standard clustering coefficient on this network without accounting for its bipartite nature.

---

#### Exercise 7. Comparative network profile

Using the `igraph` workflow from @sec-working-with-graphs-in-r, construct the following four graphs:

-   $K_5$,
-   $C_{10}$,
-   $P_{10}$,
-   one connected realisation of $G(20, 0.20)$.

For each graph, compute: order, size, $\langle k \rangle$, diameter, $c(G)$, and $\lambda_2(L)$ where applicable. Organise the results in a single table.

1.  The path $P_{10}$ and cycle $C_{10}$ have the same number of vertices but very different diameters. Express both diameters in closed form and prove your answers.
2.  Rank the four graphs by Fiedler value. Explain, in terms of graph structure, why the ranking comes out as it does.
3.  The random graph $G(20, 0.20)$ is not deterministic. Repeat the construction five times. How stable are the quantities in your table across realisations? Which quantity varies most?
4.  Write a short paragraph (5–8 sentences) identifying which of these graphs would make the most informative **null model** for comparison when, in the next chapter, we ask what a random graph typically looks like — and why.

---

## Bridge to the next chapter

This chapter has treated graphs as deterministic mathematical objects: the vertex set and edge set are given, and the task is to reason about the resulting structure. The next chapter changes this viewpoint. Instead of fixing one graph in advance, we study probability models that generate many possible graphs. The language established here — degree, adjacency matrix, connectivity, trees, Laplacian spectrum — becomes the vocabulary used to describe what a random graph typically looks like and how its properties emerge from a simple probabilistic rule.

## Concluding remarks

Graph theory provides the rigorous language on which network science depends. The definitions introduced here — graphs, adjacency matrices, degrees, paths, connectivity, trees, bipartiteness — are not merely formal conventions. They are the precise tools that make it possible to state questions unambiguously, derive results exactly, and implement algorithms correctly.

Several structural themes run through this chapter: a graph encodes only the pattern of connections; the adjacency matrix links combinatorics to linear algebra; the handshaking lemma is a universal constraint; connectivity, trees, and bipartiteness are global properties that determine much of a graph's character; and the Laplacian spectrum connects graph structure to spectral properties, with $\lambda_2$ measuring robustness.

These foundations appear in every subsequent chapter. The Erdős–Rényi [@erdos1959; @erdos1960], Watts–Strogatz [@watts1998], and Barabási–Albert [@barabasi1999] models are all built on top of the graph-theoretic language established here. Before asking how likely a graph with certain properties is, one must be able to describe those properties precisely — and that is what this chapter has provided.
