---
title: "Graph Theory"
---

```{r}
#| label: chapter-setup-graphtheory
#| include: false

setup_candidates <- c("R/book-setup.R", "../R/book-setup.R")
setup_file <- setup_candidates[file.exists(setup_candidates)][1]

if (is.na(setup_file)) {
  stop("Could not locate R/book-setup.R", call. = FALSE)
}

source(setup_file, local = knitr::knit_global())

ns_book_setup(
  preset = "core",
  seed = 42
)
```


## Introduction

Network science rests on a mathematical foundation: **graph theory**. Before studying random graphs, small worlds, or scale-free networks, it is essential to establish the formal language used to describe networks. That language is graph theory.

Graph theory is a branch of discrete mathematics that studies collections of objects and the pairwise connections between them. Its origins trace back to Leonhard Euler's 1736 solution to the *Königsberg bridge problem*, in which the question of whether a particular city could be traversed by crossing each of its seven bridges exactly once was reformulated as a purely combinatorial question about a drawing made of vertices and edges. Euler's insight — that only the pattern of connections, not the physical geography, determines the answer — is the founding idea of graph theory.

In the context of network science, graph theory provides:

-   a precise vocabulary for describing structures,
-   algebraic representations that enable computation,
-   combinatorial results that govern global properties,
-   and a rigorous basis for the probabilistic models studied in later chapters.

## Learning goals

By the end of this chapter, it should be possible to:

-   define a graph, its vertices and edges, and give the standard notation
-   distinguish between undirected, directed, weighted, and multigraphs
-   define degree, degree sequence, and prove the handshaking lemma
-   construct the adjacency matrix and the Laplacian matrix
-   define walks, paths, trails, and cycles precisely
-   define connectivity, connected components, cut vertices, and bridges
-   define trees, spanning trees, and identify their key properties
-   identify bipartite graphs and apply König's criterion
-   describe standard graph families and recognise them
-   use `igraph` in `R` to construct and analyse graphs

## What is a graph?

### Formal definition

::: {#def-graph}
A **graph** $G$ is an ordered pair

$$
G = (V, E)
$$

where $V$ is a finite non-empty set called the **vertex set** (or **node set**), and $E$ is a set of unordered pairs $\{u, v\}$ with $u, v \in V$ and $u \neq v$, called the **edge set**.
:::

The number of vertices is denoted by $N = |V|$ and is called the **order** of the graph. The number of edges is denoted by $M = |E|$ and is called the **size** of the graph.

Each element of $E$ is an unordered pair $\{u, v\}$, also written as $uv$ or $(u,v)$ when direction is irrelevant. The vertices $u$ and $v$ are said to be **adjacent** or **neighbors**, and the edge $uv$ is said to be **incident** to both $u$ and $v$.

::: {.callout-note title="Convention"}
Throughout this chapter, graphs are assumed to be **simple** unless stated otherwise. A simple graph has no self-loops (edges from a vertex to itself) and no multi-edges (more than one edge between the same pair of vertices).
:::

### A motivating example

Consider a graph with vertex set $V = \{1, 2, 3, 4, 5\}$ and edge set

$$
E = \bigl\{\{1,2\},\, \{1,3\},\, \{2,4\},\, \{3,4\},\, \{4,5\}\bigr\}.
$$

This graph has order $N = 5$ and size $M = 5$. Vertices 1, 2, and 3 are each reachable from vertex 4 through paths of length one or two, while vertex 5 hangs off vertex 4.

```{r}
#| label: simple-graph-example
#| echo: false
#| fig-cap: "A simple graph on five vertices. Vertex labels indicate node identity."

g_example <- graph_from_edgelist(
  matrix(c(1,2, 1,3, 2,4, 3,4, 4,5), ncol = 2, byrow = TRUE),
  directed = FALSE
)

layout_ex <- matrix(
  c(-1, 1,   1, 1,   -1, -0.2,   1, -0.2,   0, -1.2),
  ncol = 2, byrow = TRUE
)

par(mar = c(1, 1, 2, 1))
plot(
  g_example,
  layout = layout_ex,
  vertex.size = 28,
  vertex.color = course_cols$blue,
  vertex.frame.color = "white",
  vertex.label.color = "white",
  vertex.label.cex = 1.1,
  vertex.label.font = 2,
  edge.color = course_cols$slate,
  edge.width = 2.2,
  main = "A simple graph G = (V, E)"
)
```

## Directed and undirected graphs

The definition above assumes that each edge is unordered: if $u$ and $v$ are connected, the connection runs symmetrically in both directions. Such a graph is called **undirected**.

In many real networks, connections are asymmetric. A hyperlink may point from page $u$ to page $v$ without the reverse link existing. A neuron may send signals to a target cell without receiving signals back.

::: {#def-directed-graph}
A **directed graph** (or **digraph**) is a pair $G = (V, A)$ where $V$ is the vertex set and $A$ is a set of **ordered** pairs $(u, v)$ with $u \neq v$, called **arcs** or **directed edges**.
:::

The arc $(u, v)$ goes **from** $u$ **to** $v$. The vertex $u$ is called the **tail** and $v$ is called the **head**. The arc $(u, v)$ and the arc $(v, u)$ are distinct objects in a directed graph.

```{r}
#| label: directed-vs-undirected
#| echo: false
#| fig-cap: "Left: an undirected graph where edges have no orientation. Right: a directed graph on the same vertex set where arcs carry direction."

g_undir <- graph_from_edgelist(
  matrix(c(1,2, 2,3, 3,4, 4,1, 1,3), ncol = 2, byrow = TRUE),
  directed = FALSE
)

g_dir <- graph_from_edgelist(
  matrix(c(1,2, 3,2, 3,4, 1,4, 1,3), ncol = 2, byrow = TRUE),
  directed = TRUE
)

lyt <- matrix(c(0,1, 1,0, 0,-1, -1,0), ncol = 2, byrow = TRUE)

old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 2), mar = c(1, 1, 3, 1))

plot(g_undir,
     layout = lyt,
     vertex.size = 26, vertex.color = course_cols$teal,
     vertex.frame.color = "white", vertex.label.color = "white",
     vertex.label.font = 2,
     edge.color = course_cols$slate, edge.width = 2,
     main = "Undirected graph")

plot(g_dir,
     layout = lyt,
     vertex.size = 26, vertex.color = course_cols$gold,
     vertex.frame.color = "white", vertex.label.color = "white",
     vertex.label.font = 2,
     edge.color = course_cols$warm, edge.width = 2,
     edge.arrow.size = 0.7,
     main = "Directed graph")

par(old_par)
```

### Weighted graphs

::: {#def-weighted-graph}
A **weighted graph** is a graph $G = (V, E)$ together with a weight function $w : E \to \mathbb{R}$. Each edge $e \in E$ is assigned a real-valued weight $w(e)$.
:::

Weights may represent distances, capacities, interaction strengths, or costs, depending on the application.

## Degree and the handshaking lemma

### Degree of a vertex

::: {#def-degree}
Let $G = (V, E)$ be an undirected graph. The **degree** of a vertex $v \in V$, denoted $k_v$ or $\deg(v)$, is the number of edges incident to $v$:

$$
k_v = |\{u \in V : uv \in E\}|.
$$
:::

A vertex of degree $0$ is called **isolated**. A vertex of degree $1$ is called a **leaf** or **pendant vertex**. The maximum degree in the graph is denoted $\Delta(G)$ and the minimum degree is denoted $\delta(G)$.

### Degree sequence

::: {#def-degree-sequence}
The **degree sequence** of a graph $G$ is the list of degrees of all vertices, written in non-increasing order:

$$
k_1 \geq k_2 \geq \cdots \geq k_N.
$$
:::

The degree sequence is a fundamental invariant of the graph. It captures the distribution of connections across all vertices.

### The handshaking lemma

::: {#thm-handshaking}
**Handshaking Lemma.** For any undirected graph $G = (V, E)$,

$$
\sum_{v \in V} k_v = 2M.
$$
:::

**Proof.** Each edge $\{u, v\} \in E$ contributes exactly $1$ to the degree of $u$ and exactly $1$ to the degree of $v$. Summing over all vertices therefore counts each edge exactly twice. $\square$

::: {.callout-note title="Immediate corollary"}
The number of vertices with odd degree is always even. This follows because the sum of all degrees equals $2M$, which is even, and an even sum requires an even number of odd summands.
:::

### Average degree

::: {#def-average-degree}
The **average degree** of $G$ is

$$
\langle k \rangle = \frac{1}{N} \sum_{v \in V} k_v = \frac{2M}{N}.
$$
:::

The identity $\langle k \rangle = 2M/N$ follows immediately from the handshaking lemma. It connects average degree, edge count, and graph order in one simple formula that will appear repeatedly throughout this book.

### In-degree and out-degree in directed graphs

In a directed graph, each arc contributes to the degree of two vertices in an asymmetric way.

::: {#def-indegree-outdegree}
Let $G = (V, A)$ be a directed graph and $v \in V$. The **in-degree** of $v$ is

$$
k_v^{\text{in}} = |\{u \in V : (u, v) \in A\}|
$$

and the **out-degree** of $v$ is

$$
k_v^{\text{out}} = |\{u \in V : (v, u) \in A\}|.
$$
:::

The directed analogue of the handshaking lemma states:

$$
\sum_{v \in V} k_v^{\text{in}} = \sum_{v \in V} k_v^{\text{out}} = |A|.
$$

This holds because every arc $(u, v)$ contributes exactly $1$ to the out-degree of $u$ and exactly $1$ to the in-degree of $v$.

```{r}
#| label: degree-distribution-example
#| echo: false
#| fig-cap: "Degree sequence (sorted in decreasing order) of an example graph with 30 vertices. The dashed line marks the average degree."

set.seed(42)
g_deg <- sample_gnp(n = 30, p = 0.18)
deg_vec <- degree(g_deg)
avg_deg <- mean(deg_vec)

deg_df <- data.frame(
  vertex = seq_along(deg_vec),
  degree = sort(deg_vec, decreasing = TRUE)
)

ggplot(deg_df, aes(x = vertex, y = degree)) +
  geom_col(fill = course_cols$blue, alpha = 0.8, width = 0.8) +
  geom_hline(
    yintercept = avg_deg,
    linetype = "dashed",
    color = course_cols$warm,
    linewidth = 0.9
  ) +
annotate(
  "text",
  x = 27,
  y = avg_deg + 0.55,
  label = paste0("'\u27e8' * k * '\u27e9' == ", round(avg_deg, 2)),
  parse = TRUE,
  color = course_cols$warm,
  size = 3.8,
  hjust = 1
) +
  labs(
    title = "Degree sequence (sorted)",
    x = "Vertex rank",
    y = "Degree"
  ) +
  scale_y_continuous(breaks = 0:10)
```

## Matrix representations

### The adjacency matrix

::: {#def-adjacency-matrix}
Let $G = (V, E)$ be a graph with $V = \{1, 2, \ldots, N\}$. The **adjacency matrix** of $G$ is the $N \times N$ matrix $A$ defined by

$$
A_{ij} = \begin{cases} 1 & \text{if } \{i, j\} \in E, \\ 0 & \text{otherwise.} \end{cases}
$$
:::

For an undirected graph, the adjacency matrix is **symmetric**: $A_{ij} = A_{ji}$ for all $i, j$. For a simple graph, all diagonal entries are zero: $A_{ii} = 0$.

#### Key properties of the adjacency matrix

Several immediate properties follow from the definition:

-   The row sum of row $i$ equals the degree of vertex $i$: $\sum_{j} A_{ij} = k_i$.
-   The total sum of all entries equals twice the number of edges: $\sum_{i,j} A_{ij} = 2M$.
-   The entry $(A^r)_{ij}$ of the $r$-th power of $A$ counts the number of **walks** of length $r$ from vertex $i$ to vertex $j$.

That last property is particularly powerful. It means that matrix exponentiation yields combinatorial information about paths in the graph.

### The degree matrix

::: {#def-degree-matrix}
The **degree matrix** of $G$ is the $N \times N$ diagonal matrix $D$ where $D_{ii} = k_i$ and $D_{ij} = 0$ for $i \neq j$.
:::

### The Laplacian matrix

::: {#def-laplacian}
The **Laplacian matrix** of $G$ is

$$
L = D - A.
$$
:::

The Laplacian is symmetric and positive semi-definite. Its entries are:

$$
L_{ij} = \begin{cases} k_i & \text{if } i = j, \\ -1 & \text{if } \{i, j\} \in E, \\ 0 & \text{otherwise.} \end{cases}
$$

The smallest eigenvalue of $L$ is always $0$, with the all-ones vector $\mathbf{1}$ as eigenvector. The multiplicity of the zero eigenvalue equals the number of connected components. The second smallest eigenvalue, known as the **algebraic connectivity** or **Fiedler value** $\lambda_2$, measures how well-connected the graph is: larger $\lambda_2$ implies greater robustness to edge removal.

```{r}
#| label: adjacency-matrix-visual
#| echo: false
#| fig-cap: "Left: a small graph on six vertices. Right: its adjacency matrix displayed as a heatmap. Coloured cells indicate edges (entry = 1)."

set.seed(7)
g_mat <- graph_from_edgelist(
  matrix(c(1,2, 1,3, 2,4, 3,4, 3,5, 4,6, 5,6), ncol = 2, byrow = TRUE),
  directed = FALSE
)

A <- as.matrix(as_adjacency_matrix(g_mat))
lyt_mat <- layout_with_kk(g_mat)

old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 2), mar = c(2, 2, 3, 1))

plot(g_mat,
     layout = lyt_mat,
     vertex.size = 24,
     vertex.color = course_cols$teal,
     vertex.frame.color = "white",
     vertex.label.color = "white",
     vertex.label.font = 2,
     edge.color = course_cols$slate,
     edge.width = 2,
     main = "Graph G")

N_mat <- nrow(A)
image(
  1:N_mat, 1:N_mat, A,
  col = c("white", course_cols$teal),
  xlab = "", ylab = "",
  main = "Adjacency matrix A",
  axes = FALSE
)
axis(1, at = 1:N_mat, labels = 1:N_mat, col.axis = course_cols$ink, cex.axis = 1)
axis(2, at = 1:N_mat, labels = 1:N_mat, col.axis = course_cols$ink, cex.axis = 1, las = 1)
box(col = course_cols$grid)

par(old_par)
```

## Walks, paths, and cycles

### Walk

::: {#def-walk}
A **walk** in a graph $G = (V, E)$ is a finite sequence of vertices

$$
v_0, v_1, v_2, \ldots, v_\ell
$$

such that $\{v_{i-1}, v_i\} \in E$ for every $i = 1, \ldots, \ell$. The integer $\ell$ is the **length** of the walk.
:::

A walk may repeat vertices and edges. It simply requires that consecutive vertices in the sequence are adjacent.

### Trail and path

::: {#def-trail-path}
A **trail** is a walk in which no edge is repeated. A **path** is a walk in which no vertex is repeated (and therefore no edge is repeated either).
:::

The path is the strictest concept: it visits each vertex at most once. A trail relaxes this to edges only. Euler's original Königsberg question was precisely about the existence of a **closed trail** — a trail that returns to its starting vertex, now called an **Eulerian circuit**.

### Cycle

::: {#def-cycle}
A **cycle** is a closed walk $v_0, v_1, \ldots, v_\ell = v_0$ of length $\ell \geq 3$ in which all vertices $v_0, v_1, \ldots, v_{\ell-1}$ are distinct.
:::

The **girth** of a graph is the length of its shortest cycle. A graph with no cycles is called **acyclic**.

::: {.callout-note title="Connection to matrix powers"}
Recall that $(A^r)_{ij}$ counts walks of length $r$ from $i$ to $j$. In particular, the diagonal entry $(A^3)_{ii}$ counts closed walks of length $3$ starting and ending at $i$. The number of triangles in $G$ is therefore

$$
\text{triangles}(G) = \frac{1}{6} \operatorname{tr}(A^3).
$$

The factor $1/6$ accounts for the six ways to traverse each triangle (two directions $\times$ three starting vertices).
:::

## Connectivity

### Connected graphs and components

::: {#def-connectivity}
An undirected graph $G$ is **connected** if for every pair of distinct vertices $u, v \in V$, there exists a path from $u$ to $v$.
:::

If $G$ is not connected, it decomposes into maximal connected subgraphs, called **connected components**.

::: {#def-component}
A **connected component** of $G$ is a maximal connected subgraph of $G$.
:::

The number of connected components is denoted $c(G)$. Note that $c(G) = 1$ if and only if $G$ is connected.

### Strong and weak connectivity in directed graphs

For directed graphs, two notions of connectivity arise.

::: {#def-strong-weak-connectivity}
A directed graph is **strongly connected** if for every ordered pair $(u, v)$ there exists a directed path from $u$ to $v$. It is **weakly connected** if the underlying undirected graph is connected.
:::

Strong connectivity requires reachability in both directions between every pair of vertices. Weak connectivity only requires that edge directions can be ignored.

```{r}
#| label: connected-components-plot
#| echo: false
#| fig-cap: "A disconnected graph with three connected components shown in distinct colours. Component membership is determined by reachability."

g_comp <- make_empty_graph(n = 12, directed = FALSE)
g_comp <- add_edges(g_comp, c(1,2, 2,3, 3,1,
                               4,5, 5,6, 6,7, 7,4,
                               8,9, 9,10, 10,11, 11,12, 12,8))

comp_membership <- components(g_comp)$membership
comp_cols <- c(course_cols$blue, course_cols$teal, course_cols$gold)
vertex_cols <- comp_cols[comp_membership]

par(mar = c(1, 1, 2, 1))
set.seed(10)
plot(
  g_comp,
  layout = layout_with_fr(g_comp),
  vertex.size = 22,
  vertex.color = vertex_cols,
  vertex.frame.color = "white",
  vertex.label.color = "white",
  vertex.label.font = 2,
  vertex.label.cex = 0.9,
  edge.color = course_cols$slate,
  edge.width = 2,
  main = "Three connected components"
)
legend("bottomright",
       legend = c("Component 1 (triangle)",
                  "Component 2 (4-cycle)",
                  "Component 3 (5-cycle)"),
       fill = comp_cols, border = NA,
       bty = "n", cex = 0.88, text.col = course_cols$ink)
```

### Cut vertices and bridges

Two special structural elements are important for understanding how connectivity can be disrupted.

::: {#def-cut-vertex}
A **cut vertex** (or **articulation point**) is a vertex $v$ whose removal increases the number of connected components:

$$
c(G - v) > c(G).
$$
:::

::: {#def-bridge}
A **bridge** is an edge $e$ whose removal increases the number of connected components:

$$
c(G - e) > c(G).
$$
:::

Cut vertices and bridges are the most vulnerable elements of a connected graph. Their removal disconnects the network. Identifying them has direct relevance to network robustness, which is examined in detail in Chapter 7.

```{r}
#| label: cut-vertex-bridge
#| echo: false
#| fig-cap: "A graph with a cut vertex (red) and a bridge (dashed red edge). Removing either one disconnects the graph."

g_cut <- graph_from_edgelist(
  matrix(c(1,2, 2,3, 3,1,   # triangle on 1,2,3
           3,4,              # bridge 3-4
           4,5, 5,6, 6,4),  # triangle on 4,5,6
         ncol = 2, byrow = TRUE),
  directed = FALSE
)

lyt_cut <- matrix(c(-1.2,0, -0.4,0.9, -0.4,-0.9,
                     0.4, 0,
                     1.2, 0.9, 1.2,-0.9), ncol = 2, byrow = TRUE)

cut_v <- articulation_points(g_cut)
vertex_col <- rep(course_cols$teal, vcount(g_cut))
vertex_col[cut_v] <- "#EF4444"

bridge_idx <- which(apply(as_edgelist(g_cut), 1, function(e) {
  g_tmp <- delete_edges(g_cut, get.edge.ids(g_cut, e))
  !is_connected(g_tmp)
}))
edge_col <- rep(course_cols$slate, ecount(g_cut))
edge_lty <- rep(1, ecount(g_cut))
edge_col[bridge_idx] <- "#EF4444"
edge_lty[bridge_idx] <- 2

par(mar = c(1, 1, 2.5, 1))
plot(g_cut,
     layout = lyt_cut,
     vertex.size = 26,
     vertex.color = vertex_col,
     vertex.frame.color = "white",
     vertex.label.color = "white",
     vertex.label.font = 2,
     edge.color = edge_col,
     edge.lty = edge_lty,
     edge.width = 2.5,
     main = "Cut vertex (red node) and bridge (dashed red edge)")
```

## Trees and spanning trees

### Definition of a tree

::: {#def-tree}
A **tree** is a connected acyclic graph.
:::

Trees are among the most important and well-understood graph families. Several equivalent characterizations exist.

::: {#thm-tree-equivalences}
**Theorem.** Let $G$ be a graph with $N$ vertices. The following statements are equivalent:

1.  $G$ is a tree.
2.  $G$ is connected and has exactly $N - 1$ edges.
3.  $G$ is acyclic and has exactly $N - 1$ edges.
4.  Any two distinct vertices of $G$ are connected by exactly one path.
5.  $G$ is connected, but removing any single edge disconnects it.
:::

The equivalence of these statements is a classical result. In particular, $M = N - 1$ is a necessary and sufficient condition for a connected graph to be a tree. A **forest** is an acyclic graph that is not necessarily connected; each connected component of a forest is a tree.

### Spanning trees

::: {#def-spanning-tree}
A **spanning tree** of a connected graph $G = (V, E)$ is a subgraph $T = (V, E')$ with $E' \subseteq E$ that is itself a tree — that is, it contains all $N$ vertices with exactly $N - 1$ edges, is connected, and is acyclic.
:::

Every connected graph contains at least one spanning tree. When edge weights are present, the **minimum spanning tree** is the spanning tree for which the total edge weight is minimized. The Kruskal and Prim algorithms find minimum spanning trees in $O(M \log N)$ time.

```{r}
#| label: spanning-tree-plot
#| echo: false
#| fig-cap: "Left: a connected graph on nine vertices. Right: one spanning tree of the same graph — gold edges form the spanning tree, grey edges are excluded."

set.seed(99)
g_span <- sample_gnp(n = 9, p = 0.42)
while (!is_connected(g_span)) {
  set.seed(sample(1:1000, 1))
  g_span <- sample_gnp(n = 9, p = 0.42)
}

spanning_tree <- mst(g_span)
sp_edges <- as_edgelist(spanning_tree, names = FALSE)
all_edges <- as_edgelist(g_span, names = FALSE)

is_tree_edge <- apply(all_edges, 1, function(e) {
  any(apply(sp_edges, 1, function(se) setequal(e, se)))
})

lyt_sp <- layout_with_kk(g_span)

old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 2), mar = c(1, 1, 3, 1))

plot(g_span,
     layout = lyt_sp,
     vertex.size = 22, vertex.color = course_cols$blue,
     vertex.frame.color = "white", vertex.label.color = "white",
     vertex.label.font = 2,
     edge.color = course_cols$slate, edge.width = 1.6,
     main = "Connected graph G")

edge_col <- ifelse(is_tree_edge, course_cols$gold, scales::alpha(course_cols$slate, 0.25))
edge_wid <- ifelse(is_tree_edge, 3.0, 1.0)

plot(g_span,
     layout = lyt_sp,
     vertex.size = 22, vertex.color = course_cols$blue,
     vertex.frame.color = "white", vertex.label.color = "white",
     vertex.label.font = 2,
     edge.color = edge_col, edge.width = edge_wid,
     main = "Spanning tree (gold edges)")

par(old_par)
```

## Bipartite graphs

### Definition

::: {#def-bipartite}
A graph $G = (V, E)$ is **bipartite** if the vertex set can be partitioned into two non-empty disjoint sets $V = X \cup Y$ such that every edge in $E$ has one endpoint in $X$ and one endpoint in $Y$. The partition $(X, Y)$ is called the **bipartition**.
:::

In a bipartite graph, there are no edges within $X$ and no edges within $Y$. All edges cross the bipartition. Bipartite graphs arise naturally in models of two-mode networks: actors and films, employees and projects, genes and diseases.

### König's criterion

::: {#thm-koenig}
**Theorem (König, 1916).** A graph $G$ is bipartite if and only if it contains no cycle of odd length.
:::

This criterion is both elegant and computationally useful. Bipartiteness can be detected by a simple two-coloring of the graph, which runs in $O(N + M)$ time using breadth-first search.

### Complete bipartite graph

::: {#def-complete-bipartite}
The **complete bipartite graph** $K_{m,n}$ is the bipartite graph with parts $|X| = m$ and $|Y| = n$ in which every vertex in $X$ is adjacent to every vertex in $Y$. It has $mn$ edges and its degree sequence consists of $n$ vertices of degree $m$ and $m$ vertices of degree $n$.
:::

```{r}
#| label: bipartite-plot
#| echo: false
#| fig-cap: "Left: a bipartite graph with bipartition (blue, teal). Right: the complete bipartite graph K(3,4) in which every vertex in one part connects to every vertex in the other."

g_bip <- graph_from_biadjacency_matrix(
  matrix(c(1,0,1,1, 0,1,1,0, 1,1,0,1), nrow = 3, byrow = TRUE)
)

g_k34 <- make_full_bipartite_graph(3, 4)

lyt_bip <- layout_as_bipartite(g_bip)
lyt_k34 <- layout_as_bipartite(g_k34)

bip_type <- V(g_bip)$type
k34_type <- V(g_k34)$type

old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 2), mar = c(1, 1, 3, 1))

plot(g_bip,
     layout = lyt_bip[, c(2, 1)],
     vertex.size = 20,
     vertex.color = ifelse(bip_type, course_cols$teal, course_cols$blue),
     vertex.frame.color = "white",
     vertex.label.color = "white",
     vertex.label.font = 2,
     edge.color = course_cols$slate, edge.width = 2,
     main = "Bipartite graph")

plot(g_k34,
     layout = lyt_k34[, c(2, 1)],
     vertex.size = 20,
     vertex.color = ifelse(k34_type, course_cols$teal, course_cols$blue),
     vertex.frame.color = "white",
     vertex.label.color = "white",
     vertex.label.font = 2,
     edge.color = course_cols$slate, edge.width = 2,
     main = expression(K[3*","*4]))

par(old_par)
```

## Special graph families

Several families appear repeatedly in network science as benchmarks and building blocks.

### Complete graph

::: {#def-complete-graph}
The **complete graph** $K_N$ is the simple graph on $N$ vertices in which every pair of distinct vertices is adjacent. It has

$$
M = \binom{N}{2} = \frac{N(N-1)}{2}
$$

edges and every vertex has degree $N - 1$.
:::

### Cycle and path graphs

::: {#def-cycle-graph}
The **cycle graph** $C_N$ is the graph consisting of a single cycle through all $N$ vertices. Each vertex has degree $2$ and the graph has $N$ edges.
:::

::: {#def-path-graph}
The **path graph** $P_N$ is the graph on $N$ vertices forming a single path. The two endpoints have degree $1$, all interior vertices have degree $2$, and the graph has $N - 1$ edges.
:::

### Regular graphs

::: {#def-regular}
A graph is $k$-regular if every vertex has exactly degree $k$.
:::

The complete graph $K_N$ is $(N-1)$-regular. The cycle graph $C_N$ is $2$-regular. The Petersen graph is a famous 3-regular graph used as a counterexample in many graph-theoretic conjectures.

```{r}
#| label: special-graphs-plot
#| echo: false
#| fig-cap: "Four special graph families: the complete graph K5, the cycle C6, a 3-regular (cubic) graph on 10 vertices, and the path P6."

g_k5  <- make_full_graph(5)
g_c6  <- make_ring(6)
g_cub <- sample_k_regular(10, 3)
g_p6  <- make_lattice(c(6), circular = FALSE)

old_par <- par(no.readonly = TRUE)
par(mfrow = c(2, 2), mar = c(1, 1, 3, 1))

plot(g_k5, layout = layout_in_circle(g_k5),
     vertex.size = 22, vertex.color = course_cols$blue,
     vertex.frame.color = "white", vertex.label = NA,
     edge.color = course_cols$slate, edge.width = 1.8,
     main = expression(K[5] ~ "(complete)"))

plot(g_c6, layout = layout_in_circle(g_c6),
     vertex.size = 22, vertex.color = course_cols$teal,
     vertex.frame.color = "white", vertex.label = NA,
     edge.color = course_cols$slate, edge.width = 1.8,
     main = expression(C[6] ~ "(cycle)"))

plot(g_cub, layout = layout_with_fr(g_cub),
     vertex.size = 18, vertex.color = course_cols$gold,
     vertex.frame.color = "white", vertex.label = NA,
     edge.color = course_cols$slate, edge.width = 1.8,
     main = "3-regular (cubic) graph")

plot(g_p6, layout = matrix(c(1:6, rep(0,6)), ncol=2),
     vertex.size = 22, vertex.color = course_cols$warm,
     vertex.frame.color = "white", vertex.label = NA,
     edge.color = course_cols$slate, edge.width = 1.8,
     main = expression(P[6] ~ "(path)"))

par(old_par)
```

## Subgraphs and graph isomorphism

### Subgraph

::: {#def-subgraph}
A graph $H = (V', E')$ is a **subgraph** of $G = (V, E)$ if $V' \subseteq V$ and $E' \subseteq E$. It is an **induced subgraph** on $V' \subseteq V$ if $E'$ contains exactly all edges of $G$ with both endpoints in $V'$.
:::

The induced subgraph on a vertex set $S$ is written $G[S]$.

### Graph isomorphism

::: {#def-isomorphism}
Two graphs $G = (V_1, E_1)$ and $H = (V_2, E_2)$ are **isomorphic**, written $G \cong H$, if there exists a bijection $\phi : V_1 \to V_2$ such that

$$
\{u, v\} \in E_1 \iff \{\phi(u), \phi(v)\} \in E_2.
$$

The function $\phi$ is called a **graph isomorphism**.
:::

Isomorphic graphs are structurally identical: one is simply a relabeling of the other. Isomorphism captures the idea that only the pattern of connections — not the labels on vertices — defines a graph's structure.

A necessary condition for isomorphism is that two graphs have the same degree sequence. This condition is easy to check but is not sufficient: non-isomorphic graphs can share the same degree sequence.

## Working with graphs in R

The `igraph` package provides a comprehensive toolkit for constructing, manipulating, and analysing graphs in R.

### Constructing graphs

```{r}
#| label: igraph-construction
#| echo: true

library(igraph)

# From an edge list
g1 <- graph_from_edgelist(
  matrix(c(1,2, 1,3, 2,3, 3,4, 4,5), ncol = 2, byrow = TRUE),
  directed = FALSE
)

# From an adjacency matrix
A_mat <- matrix(c(0,1,1,0,
                  1,0,1,1,
                  1,1,0,0,
                  0,1,0,0), nrow = 4)
g2 <- graph_from_adjacency_matrix(A_mat, mode = "undirected")

# Special constructors
g_cycle  <- make_ring(8)
g_full   <- make_full_graph(6)
g_random <- sample_gnp(n = 20, p = 0.20, directed = FALSE)
```

### Basic graph attributes

```{r}
#| label: igraph-attributes
#| echo: true

cat("Order (vertices):     ", vcount(g1), "\n")
cat("Size (edges):         ", ecount(g1), "\n")
cat("Degree sequence:      ",
    paste(sort(degree(g1), decreasing = TRUE), collapse = " "), "\n")
cat("Average degree:       ", round(mean(degree(g1)), 3), "\n")
cat("Is connected?:        ", is_connected(g1), "\n")
cat("Number of components: ", components(g1)$no, "\n")
cat("Diameter:             ", diameter(g1), "\n")
cat("Girth:                ", girth(g1)$girth, "\n")
```

### Adjacency and Laplacian matrices

```{r}
#| label: igraph-matrices
#| echo: true

A_out  <- as.matrix(as_adjacency_matrix(g1))
D_out  <- diag(rowSums(A_out))
L_out  <- D_out - A_out
eigs   <- round(sort(eigen(L_out)$values), 6)

cat("Adjacency matrix:\n")
print(A_out)

cat("\nLaplacian eigenvalues:\n")
print(eigs)

cat("\nFiedler value (algebraic connectivity):", eigs[2], "\n")
```

The smallest eigenvalue is $0$ (the graph is connected, so there is exactly one zero eigenvalue). The second-smallest eigenvalue — the Fiedler value — quantifies algebraic connectivity: a larger Fiedler value indicates a more robustly connected graph.

### Triangle counting via matrix trace

```{r}
#| label: triangle-count
#| echo: true

A3    <- A_out %*% A_out %*% A_out
n_tri <- sum(diag(A3)) / 6
cat("Number of triangles (via tr(A^3)/6):", n_tri, "\n")
cat("Number of triangles (igraph)        :", sum(count_triangles(g1)) / 3, "\n")
```

## Summary of key definitions

The table below collects the most important quantities introduced in this chapter.

```{r}
#| label: summary-table
#| echo: false
#| tbl-cap: "Core graph-theoretic quantities, their notation, and formulas."

summary_df <- data.frame(
  Symbol   = c("$N = \\lvert V \\rvert$", "$M = \\lvert E \\rvert$", "$k_v$",
               "$\\langle k \\rangle$", "$A$", "$L$",
               "$d(u,v)$", "$D(G)$", "$c(G)$", "$\\lambda_2(L)$"),
  Quantity = c("Order", "Size", "Degree of $v$",
               "Average degree", "Adjacency matrix", "Laplacian matrix",
               "Distance", "Diameter", "Component count", "Fiedler value"),
  Formula  = c("$\\lvert V \\rvert$", "$\\lvert E \\rvert$", "$\\lvert \\{u: uv\\in E\\} \\rvert$",
               "$2M/N$", "$A_{ij}=\\mathbf{1}[ij\\in E]$", "$D-A$",
               "Length of shortest $u$-$v$ path",
               "$\\max_{u,v} d(u,v)$",
               "Maximal connected pieces",
               "2nd smallest eigenvalue of $L$")
)

fmt <- if (knitr::is_latex_output()) "latex" else "pipe"

knitr::kable(
  summary_df,
  format = fmt,
  align = c("c","l","l"),
  escape = FALSE,
  booktabs = knitr::is_latex_output()
)
```

## Exercises

------------------------------------------------------------------------

### Part I. Computational and derivation exercises

#### Exercise 1. Handshaking lemma

Let $G$ be an undirected simple graph with $N = 8$ vertices and degree sequence $(5, 4, 4, 3, 3, 2, 1, 0)$.

1.  Verify that the degree sequence satisfies the handshaking lemma.
2.  Determine the number of edges $M$.
3.  Compute the average degree $\langle k \rangle$.

------------------------------------------------------------------------

#### Exercise 2. Adjacency matrix

Let $G$ have vertex set $\{1,2,3,4\}$ and edge set $\{\{1,2\},\{1,3\},\{2,3\},\{3,4\}\}$.

1.  Write the adjacency matrix $A$ of $G$.
2.  Compute $A^2$ by hand.
3.  Identify what the diagonal entries of $A^2$ represent in terms of walks.
4.  Compute the number of triangles using $\operatorname{tr}(A^3)$ and verify with `igraph`.

------------------------------------------------------------------------

#### Exercise 3. Laplacian eigenvalues

For the graph in Exercise 2:

1.  Construct the degree matrix $D$ and the Laplacian $L = D - A$.
2.  Compute the eigenvalues of $L$.
3.  Verify that the smallest eigenvalue is $0$ with eigenvector $\mathbf{1}$.
4.  Interpret the second-smallest eigenvalue.

------------------------------------------------------------------------

#### Exercise 4. Trees

1.  Prove that any tree on $N$ vertices has exactly $N - 1$ edges.
2.  How many distinct labeled trees exist on $N = 4$ vertices? Draw them all.
3.  Give an example of a connected graph that is not a tree, and remove one edge to obtain a spanning tree.

------------------------------------------------------------------------

#### Exercise 5. Bipartite graphs

1.  Draw $K_{2,3}$ and compute its number of edges.
2.  Verify using König's criterion that $K_{2,3}$ contains no odd cycle.
3.  Show that the cycle graph $C_5$ is **not** bipartite.
4.  For which values of $N$ is the cycle graph $C_N$ bipartite?

------------------------------------------------------------------------

#### Exercise 6. Cut vertices and bridges

For a path graph $P_N$ with $N \geq 3$:

1.  Identify all cut vertices.
2.  Identify all bridges.
3.  What changes if $P_N$ is replaced by the cycle $C_N$?
4.  Argue that any edge of a tree is a bridge.

------------------------------------------------------------------------

### Part II. Conceptual and analytical exercises

#### Exercise 7. Degree sequence realizability

A sequence of non-negative integers $(d_1, d_2, \ldots, d_N)$ is called **graphical** if it is the degree sequence of some simple graph.

1.  Show that $(3, 3, 2, 2, 1, 1)$ is graphical by constructing a graph.
2.  Show that $(4, 3, 2, 2, 1)$ is **not** graphical using the handshaking lemma.
3.  State the Erdős--Gallai theorem and apply it to check whether $(3, 3, 2, 2, 2)$ is graphical.

------------------------------------------------------------------------

#### Exercise 8. Isomorphism

1.  Determine whether the cycle $C_6$ and the graph consisting of two disjoint triangles are isomorphic. Use degree sequences and triangle counts as invariants.
2.  Construct an explicit isomorphism or prove that none exists.

------------------------------------------------------------------------

#### Exercise 9. Powers of the adjacency matrix

Let $G = C_4$ (the 4-cycle with vertices arranged cyclically).

1.  Write the adjacency matrix $A$.
2.  Compute $A^2$ and $A^3$.
3.  Use the entries of $A^2$ to identify all walks of length $2$ between non-adjacent pairs.
4.  Explain why $(A^r)_{ij}$ counts walks, not paths.

------------------------------------------------------------------------

#### Exercise 10. Regular graphs

1.  Show that in a $k$-regular graph with $N$ vertices, the number of edges is $M = kN/2$.
2.  Deduce that $kN$ must be even. State a condition on $k$ and $N$ guaranteeing this.
3.  Show that the eigenvalues of the adjacency matrix of a $k$-regular graph all lie in the interval $[-k, k]$, with $k$ itself always an eigenvalue.

------------------------------------------------------------------------

### Part III. Simulation and coding exercises

#### Exercise 11. The Petersen graph

Using `igraph`:

1.  Construct the Petersen graph with `graph.famous("Petersen")`.
2.  Report its order, size, degree sequence, diameter, girth, and clustering coefficient.
3.  Plot it with vertex size proportional to degree.
4.  Compute its Laplacian eigenvalues. Verify that the Fiedler value is positive.

------------------------------------------------------------------------

#### Exercise 12. Exploring graph families

For each of $K_5$, $C_{10}$, $P_8$, and a random $G(20, 0.25)$ graph:

1.  Compute the degree sequence, average degree, diameter, and clustering coefficient.
2.  Determine whether the graph is bipartite using `is_bipartite()`.
3.  Find a spanning tree with `mst()` and plot it alongside the original graph.

------------------------------------------------------------------------

#### Exercise 13. Adjacency matrix powers

Write an R function that, given a graph $G$ and an integer $r$, returns the matrix $A^r$ and uses it to count:

1.  All walks of length 3 in $C_5$.
2.  The number of triangles in a random $G(15, 0.3)$ graph.

Verify both results using `igraph`'s built-in functions.

------------------------------------------------------------------------

#### Exercise 14. Detecting cut vertices

Write an R function that:

1.  Generates a random connected graph on $N$ vertices.
2.  Identifies all cut vertices using `articulation_points()`.
3.  Visualizes the graph with cut vertices highlighted in a contrasting color.
4.  Repeats this across a grid of values of $p$ from $0.1$ to $0.6$ and plots how the mean number of cut vertices changes with $p$.

------------------------------------------------------------------------

#### Exercise 15. Minimum spanning tree on a weighted graph

Generate a random graph on 12 vertices. Assign edge weights sampled uniformly from $[1, 10]$.

1.  Find the minimum spanning tree using `mst()`.
2.  Compute and report the total weight of the MST versus the full graph.
3.  Visualize the MST edges highlighted on the full graph.
4.  Repeat with weights from an exponential distribution and compare.

------------------------------------------------------------------------

## Suggested mini-project

A strong mini-project for this chapter is the following:

> Investigate graph-theoretic properties across several graph families and at least one real-world network.

A complete submission should include:

-   construction and visualization of at least four graph families,
-   computation of key quantities: degree sequence, diameter, clustering coefficient, algebraic connectivity,
-   analysis of at least one real network from the `igraphdata` package,
-   comparison of the real network's properties against the theoretical benchmarks,
-   a discussion of which graph-theoretic quantities are most informative for characterizing the network's structure.

## Concluding remarks

Graph theory provides the rigorous language that network science depends on. The definitions introduced in this chapter — graphs, adjacency matrices, degrees, paths, connectivity, trees — are not merely formal conventions. They are the precise tools that make it possible to state questions clearly, derive results exactly, and implement algorithms efficiently.

Several key ideas run through this chapter:

-   a graph encodes only the pattern of connections, not geometric or physical information,
-   the adjacency matrix links combinatorial structure to linear algebra,
-   the handshaking lemma is a universal constraint on any graph,
-   connectivity, trees, and bipartiteness are global properties that determine much of a graph's character,
-   the Laplacian matrix connects graph structure to spectral properties, with the Fiedler value measuring robustness.

These foundations will be used throughout all subsequent chapters. The Erdős--Rényi model, the Watts--Strogatz model, and the Barabási--Albert model are all built on top of the graph-theoretic language established here. Before asking *how likely* a graph with certain properties is, one must first be able to describe those properties precisely — and that is what this chapter has provided.
