---
title: "Random Graphs"
subtitle: "Probability models for network structure — the Erdős–Rényi ensemble"
bibliography: ../references.bib
csl: ../apa.csl
---

```{r}
#| label: chapter-setup
#| include: false

setup_candidates <- c("R/book-setup.R", "../R/book-setup.R")
setup_file <- setup_candidates[file.exists(setup_candidates)][1]

if (is.na(setup_file)) {
  stop("Could not locate R/book-setup.R", call. = FALSE)
}

source(setup_file, local = knitr::knit_global())

ns_book_setup(
  preset = "random",
  seed = 1234
)
```

## Introduction

Chapter 2 treated graphs as fixed mathematical objects: the vertex set and edge set were given, and the task was to reason about the resulting structure. This chapter changes that viewpoint. Instead of specifying one graph in advance, we define a **probability distribution over graphs** — a rule that assigns a probability to every possible graph on $N$ labeled vertices. The object of study is no longer a single network but an entire **ensemble** of possible networks, each realised with a specified probability.

This change of perspective is essential for network science. Real networks are not uniquely determined by a few parameters; they reflect historical, stochastic, and context-dependent processes. The ensemble viewpoint lets us ask not *what is the structure of this specific graph?* but *what structure does a graph generated by this rule typically have?* Answering this question requires expected values, distributions, and limit arguments — the tools of probability theory applied to graph-theoretic quantities.

The Erdős–Rényi models, introduced independently by @erdos1959 and @gilbert1959, are the simplest and most analytically tractable random graph models. The subsequent paper @erdos1960 established the foundational theory of phase transitions in random graphs, including the emergence of a giant component. These models serve as the primary null model in network science: if a real network's properties differ substantially from those of an Erdős–Rényi graph with the same size and density, something structurally non-trivial is present [@newman2010; @barabasi2016].

The rigorous mathematical theory of random graphs is developed in @bollobas2001book. For the network science perspective, the standard references are @newman2010 [Ch. 12--13] and @barabasi2016 [Ch. 3]. The role of the Erdős–Rényi model as a null model for community detection and other structural inference problems is discussed in @fortunato2010.

## Learning goals

By the end of this chapter, a student should be able to:

-   define a graph ensemble and distinguish $G(N,L)$ from $G(N,p)$,
-   derive the expected average degree in both models from first principles,
-   derive the exact degree distribution of $G(N,p)$ and identify its parameters,
-   carry out the Poisson limit argument in the sparse regime,
-   prove that $\mathbb{E}[C_i] = p$ in $G(N,p)$ and interpret the result,
-   connect all theoretical results to simulation in `R` and critically evaluate the fit.

## Graph ensembles

::: {#def-ensemble}
A **graph ensemble** on $N$ labeled vertices is a probability space $(\mathcal{G}_N, \mathcal{F}, \mathbb{P})$ where $\mathcal{G}_N$ is the set of all simple graphs on vertex set $\{1, \ldots, N\}$, $\mathcal{F}$ is a $\sigma$-algebra on $\mathcal{G}_N$, and $\mathbb{P}$ is a probability measure assigning $P(G) \geq 0$ to each graph $G \in \mathcal{G}_N$ with $\sum_G P(G) = 1$.
:::

In an ensemble, graph-theoretic quantities — degree, diameter, number of triangles — become **random variables**. The central questions shift from computing exact values to computing expectations, variances, and distributional limits.

## The Erdős–Rényi models

### The model $G(N,L)$

::: {#def-gnl}
The ensemble $G(N,L)$ consists of all simple graphs on $N$ labeled vertices with exactly $L$ edges, each assigned equal probability [@erdos1959].
:::

The total number of possible undirected edges on $N$ vertices is

$$
K = \binom{N}{2}.
$$

The number of labeled simple graphs with exactly $L$ edges is $\binom{K}{L}$, so the probability of any single graph $G$ in $G(N,L)$ is

$$
P(G) = \frac{1}{\binom{K}{L}}.
$$

The randomness in $G(N,L)$ lies entirely in **which** $L$ edges are chosen, not in how many. Every realisation of the model has the same number of edges.

#### Example: the ensemble $G(3,2)$

For $N = 3$: $K = \binom{3}{2} = 3$ possible edges. Choosing exactly $L = 2$ gives $\binom{3}{2} = 3$ distinct labeled graphs, each with probability $\frac{1}{3}$.

```{r}
#| label: g32-example-graphs
#| echo: false
#| fig-cap: "The three labeled graphs in the ensemble G(3,2). Each occurs with probability 1/3."

layout_triangle <- matrix(
  c(0, 1, -0.95, -0.85, 0.95, -0.85),
  ncol = 2, byrow = TRUE
)

g_32_a <- graph_from_edgelist(matrix(c("1","2","1","3"), ncol=2, byrow=TRUE), directed=FALSE)
g_32_b <- graph_from_edgelist(matrix(c("1","2","2","3"), ncol=2, byrow=TRUE), directed=FALSE)
g_32_c <- graph_from_edgelist(matrix(c("1","3","2","3"), ncol=2, byrow=TRUE), directed=FALSE)

g32_list   <- list(g_32_a, g_32_b, g_32_c)
g32_titles <- c("Edges: {1,2},{1,3}", "Edges: {1,2},{2,3}", "Edges: {1,3},{2,3}")

old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 3), mar = c(1, 1, 3, 1))
for (i in seq_along(g32_list)) {
  g_tmp <- g32_list[[i]]
  V(g_tmp)$label <- V(g_tmp)$name
  plot(g_tmp, layout = layout_triangle,
       vertex.size = 32, vertex.color = course_cols$teal,
       vertex.frame.color = "white", vertex.label.color = "white",
       vertex.label.cex = 1.05, edge.color = course_cols$blue,
       edge.width = 2.8, main = g32_titles[i])
}
par(old_par)
```

### The model $G(N,p)$

::: {#def-gnp}
The ensemble $G(N,p)$ on $N$ labeled vertices includes each of the $K = \binom{N}{2}$ possible edges independently with probability $p \in [0,1]$ [@gilbert1959].
:::

A graph $G$ with exactly $L$ edges occurs with probability

$$
P(G) = p^L (1-p)^{K-L}.
$$

Unlike $G(N,L)$, the number of edges is not fixed — it is a random variable. Two graphs with the same number of edges are equally probable; two graphs with different numbers of edges generally are not.

#### Example: the ensemble $G(3,p)$

For $N = 3$, there are $2^3 = 8$ possible graphs (each of the three edges is independently present or absent). Their probabilities depend only on $L$:

| $L$ | Number of graphs | Probability per graph |
|:---:|:---:|:---:|
| $0$ | $1$ | $(1-p)^3$ |
| $1$ | $3$ | $p(1-p)^2$ |
| $2$ | $3$ | $p^2(1-p)$ |
| $3$ | $1$ | $p^3$ |

```{r}
#| label: g3p-example-graphs
#| echo: false
#| fig-cap: "The eight labeled graphs in G(3,p), grouped by edge count L."

possible_edges <- matrix(c("1","2","1","3","2","3"), ncol=2, byrow=TRUE)

make_g3p_graph <- function(bits) {
  chosen <- possible_edges[as.logical(bits), , drop=FALSE]
  if (nrow(chosen) == 0) {
    g <- make_empty_graph(n=3, directed=FALSE)
    V(g)$name <- c("1","2","3")
  } else {
    g <- graph_from_edgelist(chosen, directed=FALSE)
    missing_nodes <- setdiff(c("1","2","3"), V(g)$name)
    if (length(missing_nodes) > 0)
      g <- add_vertices(g, nv=length(missing_nodes), name=missing_nodes)
  }
  induced_subgraph(g, vids=c("1","2","3"))
}

bit_patterns <- expand_grid(e12=c(0,1), e13=c(0,1), e23=c(0,1)) |>
  mutate(L = e12+e13+e23,
         edge_label = pmap_chr(list(e12,e13,e23), function(e12,e13,e23) {
           labs <- c()
           if (e12==1) labs <- c(labs,"(1,2)")
           if (e13==1) labs <- c(labs,"(1,3)")
           if (e23==1) labs <- c(labs,"(2,3)")
           if (length(labs)==0) return("\u2205")
           paste(labs, collapse=", ")
         }))

g3p_list   <- pmap(list(bit_patterns$e12,bit_patterns$e13,bit_patterns$e23),
                   function(e12,e13,e23) make_g3p_graph(c(e12,e13,e23)))
g3p_titles <- paste0("L = ", bit_patterns$L)

old_par <- par(no.readonly = TRUE)
par(mfrow=c(2,4), mar=c(1,1,3,1))
for (i in seq_along(g3p_list)) {
  g_tmp <- g3p_list[[i]]
  V(g_tmp)$label <- V(g_tmp)$name
  plot(g_tmp, layout=layout_triangle,
       vertex.size=28, vertex.color=course_cols$teal,
       vertex.frame.color="white", vertex.label.color="white",
       vertex.label.cex=1, edge.color=course_cols$blue,
       edge.width=2.2, main=g3p_titles[i])
}
par(old_par)
```

### Comparing the two models

::: {.callout-important title="Key distinction"}
In $G(N,L)$, every graph has exactly $L$ edges and $P(G) = 1/\binom{K}{L}$ — a uniform distribution on a restricted set.

In $G(N,p)$, every graph on $K$ possible edges is assigned a probability, but graphs with different numbers of edges have different probabilities. Only within the same edge count are graphs equiprobable.

The two models are related: $G(N,p)$ with $p = L/K$ has expected edge count $\mathbb{E}[L] = pK$, which matches the fixed $L$ of $G(N,L)$. But they are not identical — in $G(N,p)$ the edge count fluctuates around that expectation.
:::

## Expected average degree

The **average degree** of any graph with $N$ vertices and $L$ edges is

$$
\langle k \rangle = \frac{2L}{N},
$$

a consequence of the handshaking lemma from Chapter 2. In a random graph ensemble, $L$ may itself be random, so the quantity of interest is $\mathbb{E}[\langle k \rangle]$.

Since $N$ is fixed,

$$
\mathbb{E}[\langle k \rangle] = \frac{2\,\mathbb{E}[L]}{N}.
$$

### In $G(N,L)$

Every graph has exactly $L$ edges, so $\mathbb{E}[L] = L$ and

$$
\mathbb{E}[\langle k \rangle] = \frac{2L}{N}.
$$

The average degree is not merely expected to be $2L/N$ — it equals $2L/N$ in every realisation of the model.

### In $G(N,p)$

Each of the $K = \binom{N}{2}$ possible edges is included independently with probability $p$, so

$$
L \sim \mathrm{Binomial}\!\left(\binom{N}{2},\, p\right),
\qquad
\mathbb{E}[L] = p\binom{N}{2} = \frac{pN(N-1)}{2}.
$$

Substituting:

$$
\mathbb{E}[\langle k \rangle]
= \frac{2}{N} \cdot \frac{pN(N-1)}{2}
= p(N-1).
$$

A direct route to the same result: fix any vertex $v$. It has $N-1$ potential neighbors; each is connected to $v$ independently with probability $p$, so

$$
\mathbb{E}[k_v] = p(N-1).
$$

Since all vertices are statistically equivalent, this equals $\mathbb{E}[\langle k \rangle]$.

::: {.callout-note title="Key formula"}
$$
\mathbb{E}[\langle k \rangle] = p(N-1) \qquad \text{in } G(N,p).
$$

Equivalently, $p = \mathbb{E}[\langle k \rangle]/(N-1)$: to target a specific mean degree in a large network, set $p$ accordingly.
:::

```{r}
#| label: degree-distribution-simulation
#| fig-cap: "Observed degree frequencies in one realisation of G(80, 0.08) compared with the theoretical binomial distribution."

n <- 80
p <- 0.08

g <- sample_gnp(n=n, p=p, directed=FALSE, loops=FALSE)
obs_deg <- degree(g)

obs_tbl <- tibble(k=obs_deg) |>
  count(k, name="observed") |>
  complete(k=0:(n-1), fill=list(observed=0)) |>
  mutate(observed_prob = observed / sum(observed))

theory_tbl <- tibble(k=0:(n-1)) |>
  mutate(theory_prob = dbinom(k, size=n-1, prob=p))

plot_tbl <- left_join(obs_tbl, theory_tbl, by="k")

ggplot(plot_tbl, aes(x=k)) +
  geom_col(aes(y=observed_prob), fill=course_cols$blue, alpha=0.5, width=0.82) +
  geom_line(aes(y=theory_prob), color=course_cols$gold, linewidth=1.1) +
  geom_point(aes(y=theory_prob), color=course_cols$teal, size=2.2) +
  labs(title="Observed and theoretical degree distribution",
       subtitle=sprintf("Single realisation of G(%d, %.2f)", n, p),
       x="Degree k", y="Probability")
```

## Degree distribution in $G(N,p)$

Fix a vertex $v$. It has $N-1$ potential neighbors. Each potential edge is present independently with probability $p$ and absent with probability $1-p$. The degree $k_v$ is therefore the count of successes in $N-1$ independent Bernoulli$(p)$ trials:

$$
k_v \sim \mathrm{Binomial}(N-1,\, p).
$$

The probability that $v$ has degree exactly $k$ is

$$
P(k) = \binom{N-1}{k}\, p^k\, (1-p)^{N-1-k}, \qquad k = 0, 1, \ldots, N-1.
$$

The three factors have clear interpretations:

- $\binom{N-1}{k}$: the number of ways to choose which $k$ of the $N-1$ potential neighbors are connected,
- $p^k$: the probability that those $k$ edges are present,
- $(1-p)^{N-1-k}$: the probability that the remaining $N-1-k$ edges are absent.

The mean and variance of this distribution are

$$
\mathbb{E}[k] = p(N-1), \qquad \mathrm{Var}(k) = p(1-p)(N-1).
$$

The mean recovers the expected average degree derived above. The variance quantifies fluctuations around that mean within a single graph.

::: {.callout-important title="Summary"}
In $G(N,p)$, the degree of any vertex follows a $\mathrm{Binomial}(N-1, p)$ distribution. The degree distribution of the model is therefore

$$
P(k) = \binom{N-1}{k} p^k (1-p)^{N-1-k}.
$$
:::

## Poisson approximation in the sparse regime

The binomial degree distribution is exact. In many network science settings, however, attention focuses on large graphs in the **sparse regime**, where

$$
N \to \infty, \quad p \to 0, \quad \lambda := p(N-1) \text{ remains finite.}
$$

Sparsity means the average degree stays bounded even as the network grows — a feature shared by many real-world networks.

### Derivation of the Poisson limit

Starting from the exact binomial probability:

$$
P(k) = \binom{N-1}{k}\, p^k\, (1-p)^{N-1-k}.
$$

Expand the binomial coefficient and substitute $p = \lambda/(N-1)$:

$$
\binom{N-1}{k} = \frac{(N-1)(N-2)\cdots(N-k)}{k!}.
$$

This gives

$$
P(k) = \frac{(N-1)(N-2)\cdots(N-k)}{k!}
\cdot \left(\frac{\lambda}{N-1}\right)^k
\cdot \left(1 - \frac{\lambda}{N-1}\right)^{N-1-k}.
$$

Rearranging:

$$
P(k) = \frac{\lambda^k}{k!}
\cdot \underbrace{\frac{(N-1)(N-2)\cdots(N-k)}{(N-1)^k}}_{\to\, 1}
\cdot \underbrace{\left(1-\frac{\lambda}{N-1}\right)^{N-1}}_{\to\, e^{-\lambda}}
\cdot \underbrace{\left(1-\frac{\lambda}{N-1}\right)^{-k}}_{\to\, 1}.
$$

As $N \to \infty$ with $\lambda$ fixed:

1. The first product $\frac{(N-1)(N-2)\cdots(N-k)}{(N-1)^k} = \prod_{j=0}^{k-1}\left(1 - \frac{j}{N-1}\right) \to 1$ since each factor tends to $1$ and $k$ is fixed.
2. $\left(1 - \frac{\lambda}{N-1}\right)^{N-1} \to e^{-\lambda}$ by the standard limit $\lim_{m\to\infty}(1-x/m)^m = e^{-x}$.
3. $\left(1 - \frac{\lambda}{N-1}\right)^{-k} \to 1$ since $k$ is fixed and $\lambda/(N-1) \to 0$.

Therefore,

$$
\boxed{P(k) \;\longrightarrow\; e^{-\lambda}\frac{\lambda^k}{k!},}
$$

the Poisson distribution with mean $\lambda = p(N-1) = \mathbb{E}[k]$.

::: {.callout-note title="What the approximation requires"}
The Poisson approximation is accurate when $N$ is large and $p$ is small such that $\lambda = p(N-1)$ is of moderate size. It fails when $p$ is not small (e.g., in dense graphs), because then $\mathrm{Var}(k) = p(1-p)(N-1) \approx (1-p)\lambda \not\approx \lambda$, and the Poisson assumption of equal mean and variance breaks down.
:::

```{r}
#| label: poisson-approximation
#| fig-cap: "Binomial degree distribution and its Poisson approximation for G(500, 0.02). The two curves are nearly indistinguishable in the sparse regime."

n      <- 500
p      <- 0.02
lambda <- p * (n - 1)
k_vals <- 0:24

compare_tbl <- tibble(
  k = k_vals,
  binomial = dbinom(k_vals, size=n-1, prob=p),
  poisson  = dpois(k_vals, lambda=lambda)
) |>
  pivot_longer(cols=c(binomial, poisson),
               names_to="distribution", values_to="probability") |>
  mutate(distribution = recode(distribution,
                               binomial="Binomial", poisson="Poisson"))

ggplot(compare_tbl, aes(x=k, y=probability, color=distribution)) +
  geom_line(linewidth=1.1) +
  geom_point(size=2.2) +
  scale_color_manual(values=c("Binomial"=course_cols$gold,
                              "Poisson"=course_cols$teal)) +
  labs(title="Binomial and Poisson degree distributions",
       subtitle=sprintf("N = %d, p = %.3f, mean degree = %.2f", n, p, lambda),
       x="Degree k", y="Probability", color=NULL)
```

## Clustering coefficient

### Definition

The **local clustering coefficient** of vertex $i$ with degree $k_i$ measures the fraction of its $\binom{k_i}{2}$ possible neighbor-neighbor edges that are actually present. If $L_i$ denotes the number of edges among the neighbors of $i$, then

$$
C_i = \frac{L_i}{\binom{k_i}{2}} = \frac{2L_i}{k_i(k_i-1)}, \qquad k_i \geq 2.
$$

For $k_i \leq 1$, $C_i$ is conventionally set to $0$. The value $C_i \in [0,1]$: it equals $0$ when no two neighbors of $i$ are connected, and $1$ when all $\binom{k_i}{2}$ possible neighbor pairs are connected.

The **average clustering coefficient** of the graph is

$$
\bar{C} = \frac{1}{N}\sum_{i=1}^N C_i.
$$

```{r}
#| label: clustering-examples
#| echo: false
#| fig-cap: "Three neighborhoods of the same focal node A, each with degree 3 but different numbers of neighbor-neighbor edges. Teal edges connect A to its neighbors; gold edges connect neighbors to each other."

layout_clust <- matrix(c(0,0, -1.0,0.9, -0.2,-1.0, 1.0,0.9),
                       ncol=2, byrow=TRUE)

plot_clustering_example <- function(edges_center, edges_neighbor, title_text) {
  g <- make_empty_graph(n=4, directed=FALSE)
  V(g)$name <- c("A","B","C","D")
  all_edges <- c(edges_center, edges_neighbor)
  if (length(all_edges) > 0) g <- add_edges(g, all_edges)
  edge_pairs <- ends(g, es=E(g), names=TRUE)
  edge_cols  <- apply(edge_pairs, 1, function(x)
    if ("A" %in% x) course_cols$teal else course_cols$gold)
  edge_widths <- apply(edge_pairs, 1, function(x)
    if ("A" %in% x) 3.2 else 3.8)
  plot(g, layout=layout_clust,
       vertex.size=c(34,30,30,30),
       vertex.color=c(course_cols$gold, course_cols$blue,
                      course_cols$blue, course_cols$blue),
       vertex.frame.color="white", vertex.label.color="white",
       vertex.label.cex=1, edge.color=edge_cols,
       edge.width=edge_widths, main=title_text)
}

old_par <- par(no.readonly = TRUE)
par(mfrow=c(1,3), mar=c(1,1,3,1))

plot_clustering_example(
  edges_center   = c("A","B","A","C","A","D"),
  edges_neighbor = c(),
  title_text     = expression(L[A]==0 ~~ C[A]==0))
plot_clustering_example(
  edges_center   = c("A","B","A","C","A","D"),
  edges_neighbor = c("B","C"),
  title_text     = expression(L[A]==1 ~~ C[A]==1/3))
plot_clustering_example(
  edges_center   = c("A","B","A","C","A","D"),
  edges_neighbor = c("B","C","B","D","C","D"),
  title_text     = expression(L[A]==3 ~~ C[A]==1))

par(old_par)
```

### Expected clustering in $G(N,p)$

::: {#thm-clustering-gnp}
**Proposition.** In $G(N,p)$,

$$
\mathbb{E}[C_i] = p \quad \text{for every vertex } i \text{ with } k_i \geq 2,
$$

and consequently $\mathbb{E}[\bar{C}] = p$.
:::

**Proof.** Condition on a fixed set of $k_i$ neighbors of vertex $i$. For each pair $(u,v)$ among those neighbors, the edge $\{u,v\}$ is present independently with probability $p$ — independently of the edges connecting $u$ and $v$ to $i$, because edges in $G(N,p)$ are mutually independent. Therefore

$$
\mathbb{E}\!\left[L_i \,\big|\, k_i\right] = \binom{k_i}{2} p,
$$

which gives

$$
\mathbb{E}\!\left[C_i \,\big|\, k_i\right] = \frac{\mathbb{E}[L_i \mid k_i]}{\binom{k_i}{2}} = p.
$$

Taking expectations over $k_i$: $\mathbb{E}[C_i] = p$. Averaging over all vertices: $\mathbb{E}[\bar{C}] = p$. $\square$

::: {.callout-note title="Interpretation"}
In $G(N,p)$, clustering is entirely determined by the edge probability $p$. In sparse graphs where $p$ is small, clustering is correspondingly small. This is a direct consequence of edge independence: there is no mechanism favouring triangle closure beyond the baseline probability that any edge exists.

Many empirical social networks have clustering coefficients far exceeding $p$, even when matched to a random graph on the same $N$ and average degree [@watts1998; @newman2010]. This gap — between Erdős–Rényi clustering and observed clustering — motivates the small-world models studied in the next chapter [@watts1998; @strogatz2001].
:::

```{r}
#| label: simulation-summary

n    <- 100
p    <- 0.07
reps <- 250

sim_tbl <- map_dfr(seq_len(reps), function(i) {
  g <- sample_gnp(n=n, p=p, directed=FALSE, loops=FALSE)
  tibble(
    sim            = i,
    avg_degree     = mean(degree(g)),
    avg_clustering = transitivity(g, type="average")
  )
})

summary_tbl <- tibble(
  quantity          = c("Average degree", "Average clustering"),
  empirical_mean    = c(mean(sim_tbl$avg_degree),
                        mean(sim_tbl$avg_clustering, na.rm=TRUE)),
  theoretical_value = c(p*(n-1), p)
)

summary_tbl
```

```{r}
#| label: clustering-plot
#| fig-cap: "Distribution of average clustering across 250 realisations of G(100, 0.07). Dashed line marks the theoretical value p."

ggplot(sim_tbl, aes(x=avg_clustering)) +
  geom_histogram(bins=24, fill=course_cols$gold, alpha=0.65, color="white") +
  geom_vline(xintercept=p, color=course_cols$teal,
             linewidth=1.1, linetype=2) +
  labs(title="Average clustering across simulations",
       subtitle=sprintf("Dashed line = theoretical value p = %.2f", p),
       x="Average clustering", y="Count")
```

## Core formulas

::: {.callout-important title="Summary of key results"}
For Erdős–Rényi random graphs:

| Quantity | $G(N,L)$ | $G(N,p)$ |
|:---|:---:|:---:|
| $P(G)$ | $1/\binom{K}{L}$ | $p^L(1-p)^{K-L}$ |
| $\mathbb{E}[\langle k \rangle]$ | $2L/N$ | $p(N-1)$ |
| Degree distribution | — | $\mathrm{Bin}(N-1,p)$ |
| Poisson limit (sparse) | — | $\mathrm{Poisson}(\lambda)$, $\lambda=p(N-1)$ |
| $\mathbb{E}[\bar{C}]$ | — | $p$ |

where $K = \binom{N}{2}$.
:::

## Working with random graphs in R {#sec-working-with-random-graphs-in-r}

```{r}
#| label: er-graph-example
#| fig-cap: "One realisation of G(20, 0.12)."

n <- 20
p <- 0.12

g_er <- sample_gnp(n=n, p=p, directed=FALSE, loops=FALSE)

plot(g_er,
     layout=layout_with_fr(g_er),
     vertex.size=22, vertex.color=course_cols$teal,
     vertex.frame.color="white", vertex.label.color="white",
     edge.color=scales::alpha(course_cols$ink, 0.32),
     vertex.label.cex=0.8,
     main=sprintf("A realisation of G(%d, %.2f)", n, p))
```

```{r}
#| label: interactive-helper
#| echo: false

degree_dist_df <- function(n_value, p_value) {
  tibble(k=0:(n_value-1),
         prob=dbinom(0:(n_value-1), size=n_value-1, prob=p_value))
}

make_degree_static <- function(n_value, p_value) {
  df <- degree_dist_df(n_value, p_value)
  ggplot(df, aes(x=k, y=prob)) +
    geom_col(fill=course_cols$blue, alpha=0.45, width=0.82) +
    geom_line(color=course_cols$gold, linewidth=1.1) +
    geom_point(color=course_cols$teal, size=2.2) +
    labs(title=sprintf("Degree distribution for G(%d, %.2f)", n_value, p_value),
         subtitle=sprintf("Expected degree = p(N-1) = %.2f", p_value*(n_value-1)),
         x="Degree k", y="Probability")
}

make_degree_interactive <- function(n_value,
    p_values=c(0.04, 0.08, 0.12, 0.16, 0.20)) {
  df <- expand_grid(p=p_values, k=0:(n_value-1)) |>
    mutate(prob=dbinom(k, size=n_value-1, prob=p),
           frame_lab=paste0("p = ", sprintf("%.2f", p)),
           hover_lab=paste0("N = ",n_value,
                            "<br>p = ", sprintf("%.2f", p),
                            "<br>k = ",k,
                            "<br>P(K=k) = ",sprintf("%.4f",prob),
                            "<br>Expected degree = ",sprintf("%.2f",p*(n_value-1))))
  plot_ly(df, x=~k, y=~prob, frame=~frame_lab,
          type="scatter", mode="lines+markers",
          text=~hover_lab, hoverinfo="text",
          line=list(color=course_cols$teal, width=3),
          marker=list(color=course_cols$gold, size=7)) |>
    layout(title=list(text=paste0("Interactive degree distribution for N = ",n_value)),
           xaxis=list(title="Degree k", rangemode="tozero"),
           yaxis=list(title="Probability", rangemode="tozero"),
           plot_bgcolor="#FCFCFD", paper_bgcolor="#FCFCFD",
           showlegend=FALSE) |>
    animation_opts(frame=1100, transition=0, redraw=FALSE) |>
    animation_slider(currentvalue=list(prefix="Current setting: ")) |>
    config(displayModeBar=FALSE)
}
```

::: {.content-visible when-format="html"}

### Explore the effect of $p$ {.panel-tabset}

#### $N = 30$

```{r}
#| echo: false
if (knitr::is_html_output()) make_degree_interactive(30)
```

#### $N = 60$

```{r}
#| echo: false
if (knitr::is_html_output()) make_degree_interactive(60)
```

#### $N = 100$

```{r}
#| echo: false
if (knitr::is_html_output()) make_degree_interactive(100)
```

:::

::: {.content-visible unless-format="html"}

```{r}
#| echo: false
if (knitr::is_latex_output()) make_degree_static(60, 0.12)
```

:::

## In-class questions

1.  What distinguishes a graph ensemble from a single graph?
2.  Why does $L$ follow a binomial distribution in $G(N,p)$ but not in $G(N,L)$?
3.  What does the Poisson approximation require, and why does it fail in the dense regime?
4.  Two graphs have the same $N$ and $\mathbb{E}[\langle k \rangle]$. Must they have the same degree distribution?
5.  Why is $\mathbb{E}[\bar{C}] = p$ a consequence of edge independence, not of any structural property of $G$?
6.  A real social network has clustering $\bar{C} = 0.4$ and $p = 0.01$. What does this tell you?

## Exercises

The seven exercises below cover the full content of this chapter. They are designed to require reasoning and derivation, not reproduction of results already in the text. The first two are mathematical, the next three analytical, and the final two computational.

---

#### Exercise 1. Model structure and equiprobability

Let $G(3,2)$ and $G(3,p)$ with $p = 2/3$ both be defined on the same three labeled vertices.

1. List all graphs in $G(3,2)$ and compute $P(G)$ for each. Are all graphs with two edges in $G(3,p)$ assigned the same probability? Are they assigned the same probability as in $G(3,2)$?
2. In $G(N,p)$, two graphs $G_1$ and $G_2$ with $L_1 \neq L_2$ edges satisfy $P(G_1) \neq P(G_2)$ whenever $p \in (0,1)$. Prove this directly from the probability formula $P(G) = p^L(1-p)^{K-L}$.
3. For what value of $p$ is $G(N,p)$ the uniform distribution over **all** $2^K$ labeled simple graphs? Justify your answer.
4. In $G(N,L)$, two vertices $u$ and $v$ are connected with probability $L/K$ where $K = \binom{N}{2}$. Derive this result combinatorially. *(Hint: count the graphs in the ensemble that contain the edge $\{u,v\}$.)*

---

#### Exercise 2. Expected degree: derivation and inversion

1. Derive $\mathbb{E}[\langle k \rangle] = p(N-1)$ in $G(N,p)$ using two different arguments: first via $\mathbb{E}[L]$, then directly from the degree of a single vertex. Confirm that both give the same result.
2. Compute $\mathrm{Var}(\langle k \rangle)$ in $G(N,p)$. Show that the relative fluctuation $\mathrm{SD}(\langle k \rangle)/\mathbb{E}[\langle k \rangle]$ vanishes as $N \to \infty$ with $\lambda = p(N-1)$ fixed. What does this say about how close $\langle k \rangle$ is to its expectation in large graphs?
3. You observe a network with $N = 1000$ vertices and average degree $\langle k \rangle = 6$. If you model it as $G(N,p)$, what value of $p$ should you use? Would you expect the degree distribution to be closer to binomial or Poisson? Justify your answer.

---

#### Exercise 3. Degree distribution: properties and moments

The degree distribution in $G(N,p)$ is $P(k) = \binom{N-1}{k}p^k(1-p)^{N-1-k}$.

1. Verify that $\sum_{k=0}^{N-1} P(k) = 1$ using the binomial theorem. Show your working.
2. Compute the variance $\mathrm{Var}(k) = p(1-p)(N-1)$ from the definition $\mathrm{Var}(k) = \mathbb{E}[k^2] - (\mathbb{E}[k])^2$.
3. In the sparse regime $\lambda = p(N-1)$ fixed, the Poisson distribution has $\mathrm{Var}(k) = \lambda = \mathbb{E}[k]$. Show that $\mathrm{Var}(k) = p(1-p)(N-1) \to \lambda$ as $N \to \infty$ with $\lambda$ fixed. What property of the Poisson distribution does this reflect?
4. A student argues: *"Since both the binomial and the Poisson have the same mean $\lambda$ in the sparse regime, there is no reason to prefer one over the other."* Identify precisely what is wrong with this argument, and state one quantity for which the two distributions give different predictions.

---

#### Exercise 4. The Poisson limit

1. Carry out the Poisson limit argument from the lecture in full detail for the case $k = 2$. That is, starting from $P(2) = \binom{N-1}{2}p^2(1-p)^{N-3}$ with $p = \lambda/(N-1)$, show explicitly that $P(2) \to e^{-\lambda}\lambda^2/2$ as $N \to \infty$.
2. The argument uses three separate limit steps. State each one precisely and identify the standard limit result being applied.
3. Suppose instead of fixing $\lambda = p(N-1)$, you fix $\mu = pN$. Does the Poisson approximation still hold? If so, with which mean parameter? If not, explain what goes wrong.
4. For which values of $p$ (as a function of $N$) is the Poisson approximation *not* appropriate, and why?

---

#### Exercise 5. Clustering: proof and interpretation

1. The lecture proves $\mathbb{E}[C_i] = p$ by conditioning on the neighborhood of vertex $i$. Write out this proof in full, being explicit about which independence assumption is used at each step.
2. A student constructs a graph model in which edges are not independent: whenever edge $\{u,v\}$ is present, each of the edges $\{u,w\}$ and $\{v,w\}$ for any third vertex $w$ is present with probability $q > p$ (rather than $p$). Qualitatively, how would $\mathbb{E}[\bar{C}]$ compare to $p$ in this model? Relate your answer to real social networks.
3. In $G(N,p)$, the expected number of triangles is $\binom{N}{3}p^3$. Derive this result. Then express the expected clustering coefficient in terms of the expected number of triangles and the expected number of connected triples (paths of length 2). Are the two expressions for $\mathbb{E}[\bar{C}]$ consistent?
4. Fix $N = 100$ and vary $p$ from $0.01$ to $0.5$. Predict how $\mathbb{E}[\bar{C}]$ and $\mathbb{E}[\langle k \rangle]$ scale with $p$. Does the ratio $\mathbb{E}[\bar{C}]/\mathbb{E}[\langle k \rangle]$ depend on $N$?

---

#### Exercise 6. Comparing $G(N,L)$ and $G(N,p)$ numerically

Set $N = 50$, $L = 100$, and $p = L/\binom{N}{2}$.

1. Generate 500 independent graphs from $G(N,p)$ using `sample_gnp`. For each graph record: edge count $M$, average degree $\langle k \rangle$, and average clustering $\bar{C}$.
2. Plot the distribution of edge counts across the 500 realisations. Mark $L = 100$ and $\mathbb{E}[L] = pK$ on the plot. Comment on the spread and its relationship to $\mathrm{Var}(L) = Kp(1-p)$.
3. Generate one graph from $G(N,L)$ using `sample_gnm(N, L)`. Compare its average degree and clustering to the empirical means from part 1. Is this comparison fair? What does it reveal about the difference between the two models?
4. Compute the empirical degree distribution across all 500 graphs from $G(N,p)$. Overlay the theoretical binomial $\mathrm{Bin}(N-1, p)$ and the Poisson approximation $\mathrm{Poisson}(\lambda)$ on the same plot. At this value of $N$ and $p$, which approximation fits better? Is this what you expected?

---

#### Exercise 7. A network diagnostic: ER as a null model

A network analyst observes a real network with the following properties: $N = 200$, $M = 600$, $\bar{C} = 0.18$, and degree sequence ranging from $1$ to $25$.

1. Fit an Erdős–Rényi model $G(N,p)$ to this network by matching the observed average degree. What value of $p$ do you use?
2. Under your fitted model, compute: (a) $\mathbb{E}[\bar{C}]$, (b) the expected maximum degree, (c) the probability that a randomly chosen vertex has degree $0$.
3. The observed clustering $\bar{C} = 0.18$ is much larger than $\mathbb{E}[\bar{C}]$ under your fitted model. Simulate 200 realisations of $G(N,p)$ and plot the distribution of $\bar{C}$. Where does $0.18$ fall in this distribution? What does this tell you about the real network?
4. The observed degree sequence has maximum degree $25$. Under $G(N,p)$ with your fitted $p$, what is the probability of observing a vertex of degree $\geq 25$? Compute this analytically and verify it by simulation. What structural feature of the real network might explain a higher maximum degree than the ER model predicts?

---

## Bridge to the next chapter

The Erdős–Rényi model captures one key feature of real networks — sparse random connectivity — but fails on two others: it has low clustering ($\mathbb{E}[\bar{C}] = p \approx 0$ for sparse graphs), and its degree distribution is binomial or Poisson, decaying exponentially for large $k$. Empirical networks routinely exhibit clustering far above the ER baseline [@watts1998] and degree distributions with much heavier tails — some vertices have degrees orders of magnitude larger than the mean [@barabasi1999; @albert2002]. These two departures from the ER model motivate the next two chapters: the Watts–Strogatz model [@watts1998] introduces a mechanism for generating high clustering at low average degree, and the Barabási–Albert model [@barabasi1999] introduces preferential attachment to generate heavy-tailed degree distributions. The degree sequence approach of @molloy1995 provides a third generalisation, allowing arbitrary degree distributions to be studied within a random graph framework.

## Concluding remarks

The Erdős–Rényi model is not a realistic description of most real networks. Its value lies elsewhere: it is the simplest probability model over graphs, it is analytically exact [@erdos1959; @erdos1960; @bollobas2001book], and it provides a rigorous null hypothesis against which observed network structure can be tested [@newman2010]. The binomial degree distribution, the Poisson limit, and the clustering formula $\mathbb{E}[\bar{C}] = p$ are all consequences of a single assumption — that edges form independently. When real networks violate this assumption, as they almost always do [@watts1998; @barabasi1999; @strogatz2001], the deviation itself is informative. That is why the Erdős–Rényi model remains the essential starting point for network science.
