---
title: "Scale-Free Networks"
---

```{r}
#| label: chapter-setup-scalefree
#| include: false

library(igraph)
library(dplyr)
library(ggplot2)
library(tidyr)
library(purrr)

course_cols <- list(
  teal  = "#0F766E",
  blue  = "#2563EB",
  gold  = "#F59E0B",
  coral = "#E76F51",
  ink   = "#1F2937",
  grid  = "#E5E7EB"
)

theme_set(
  theme_minimal(base_size = 13) +
    theme(
      plot.title.position = "plot",
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_line(color = course_cols$grid, linewidth = 0.35),
      panel.grid.major.y = element_line(color = course_cols$grid, linewidth = 0.35),
      axis.title = element_text(color = course_cols$ink),
      axis.text = element_text(color = course_cols$ink),
      plot.title = element_text(face = "bold", color = course_cols$ink),
      plot.subtitle = element_text(color = "#475569"),
      legend.position = "top"
    )
)

set.seed(2040)
```

## Introduction

The Erdős--Rényi model and the Watts--Strogatz model explain important aspects of network structure, but many real systems display a feature that neither model captures naturally: the existence of a small number of highly connected vertices, usually called **hubs**.

Examples include:

-   highly cited scientific papers,
-   very popular web pages,
-   airports with exceptionally high traffic,
-   social actors with unusually many connections.

In such systems, degree is not concentrated around a typical value. Instead, the degree distribution is strongly heterogeneous. Most vertices have small degree, but a few vertices have extremely large degree. This broad distribution is one of the defining signatures of a scale-free network.

The purpose of this chapter is to study the mathematical meaning of scale-free behavior and the mechanism by which it can arise. The central model is the **Barabási--Albert model**, which combines two key ingredients:

1.  growth,
2.  preferential attachment.

Together they generate networks with heavy-tailed degree distributions and hub formation.

## Learning goals

By the end of this chapter, it should be possible to:

-   define degree heterogeneity and hub structure
-   define a power-law degree distribution
-   explain what is meant by a scale-free network
-   define preferential attachment mathematically
-   describe the Barabási--Albert model
-   explain why preferential attachment leads to hub formation
-   verify scale-free behavior numerically in `R`

## Broad degree distributions and hubs

### Degree heterogeneity

A first step toward scale-free structure is to understand that degree distributions need not be concentrated around one typical value.

In a homogeneous network, most vertices have degree values of roughly the same order, and the distribution is relatively narrow. By contrast, in a heterogeneous network, degree values span many scales.

::: {#def-degree-heterogeneity}
A network is said to exhibit **degree heterogeneity** if its degree distribution is broad, in the sense that vertices of very different degree coexist in significant numbers.
:::

This definition is qualitative, but it captures an essential structural distinction. In a heterogeneous network, one cannot meaningfully describe the graph using only a “typical” degree.

### Hubs

::: {#def-hub}
A **hub** is a vertex whose degree is substantially larger than the degrees of most other vertices in the same network.
:::

The notion of a hub is relative to the graph under consideration. A vertex of degree $20$ may be a hub in one network and completely ordinary in another.

The significance of hubs is not merely visual. They strongly influence:

-   shortest paths,
-   robustness,
-   spreading processes,
-   centralization of connectivity.

So the appearance of hubs is one of the main reasons why scale-free networks differ from more homogeneous random graphs.

### Visual contrast: homogeneous and heterogeneous degree structure

```{r}
#| label: heterogeneous-vs-homogeneous
#| echo: false
#| fig-cap: "A comparison between a homogeneous random graph and a heterogeneous preferential-attachment graph with approximately similar average degree."

g_er <- sample_gnp(n = 50, p = 0.08, directed = FALSE, loops = FALSE)
g_ba <- sample_pa(n = 50, power = 1, m = 2, directed = FALSE)

plot_degree_colored <- function(g, main_text) {
  deg <- degree(g)
  brks <- quantile(deg, probs = c(0, 0.5, 0.8, 1), na.rm = TRUE)
  brks[1] <- brks[1] - 1e-8
  cols <- cut(
    deg,
    breaks = unique(brks),
    include.lowest = TRUE,
    labels = c(course_cols$blue, course_cols$teal, course_cols$coral)[seq_len(length(unique(brks)) - 1)]
  )

  plot(
    g,
    layout = layout_with_fr(g),
    vertex.size = 8 + 2.2 * sqrt(deg + 1),
    vertex.color = as.character(cols),
    vertex.frame.color = "white",
    vertex.label = NA,
    edge.color = scales::alpha(course_cols$ink, 0.20),
    edge.width = 1.0,
    main = main_text
  )
}

old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 2), mar = c(1, 1, 3, 1))

plot_degree_colored(g_er, "Homogeneous random graph")
plot_degree_colored(g_ba, "Preferential-attachment graph")

par(old_par)
```

The comparison above illustrates the basic distinction:

-   in a homogeneous random graph, vertex degrees tend to lie in a relatively narrow range
-   in a preferential-attachment graph, a few vertices accumulate much larger degree and emerge as hubs

This does not yet prove a power law, but it motivates the need for a more precise mathematical description.

## Power-law degree distributions

### Definition of a power law

::: {#def-power-law}
A degree distribution is said to follow a **power law** if, for sufficiently large degree $k$,

$$
P(k) \sim Ck^{-\gamma},
$$

where $C>0$ is a constant and $\gamma > 0$ is the degree exponent.
:::

The notation $\sim$ means asymptotic proportionality. More precisely,

$$
\lim_{k \to \infty} \frac{P(k)}{Ck^{-\gamma}} = 1.
$$

So a power law describes the tail behavior of the degree distribution.

### Why the term “scale-free” is used

Suppose

$$
P(k) \sim Ck^{-\gamma}.
$$

Then for any scaling factor $a>0$,

$$
P(ak) \sim C(ak)^{-\gamma} = a^{-\gamma} Ck^{-\gamma}.
$$

So changing the scale of $k$ only multiplies the distribution by a constant factor. There is no preferred characteristic scale in the tail. This is the reason for the term **scale-free**.

::: {#def-scale-free-network}
A network is called **scale-free** if its degree distribution has a power-law tail over a substantial range of degrees.
:::

This definition should be interpreted with care. In applications, one usually does not expect a perfect power law over all degrees. What matters is that the tail decays approximately as a power law over a meaningful interval.

### Cumulative distribution

In empirical work it is often useful to consider the cumulative degree distribution

$$
P(K \geq k).
$$

If

$$
P(k) \sim Ck^{-\gamma},
$$

then the cumulative distribution satisfies

$$
P(K \geq k) \sim C' k^{-(\gamma-1)}
$$

for large $k$, provided $\gamma > 1$.

This representation is often smoother numerically and easier to visualize on a log-log scale.

## Preferential attachment

The central mechanism behind the Barabási--Albert model is preferential attachment.

::: {#def-preferential-attachment}
A growing network is said to evolve by **preferential attachment** if the probability that a new vertex connects to an existing vertex $i$ is proportional to the degree of $i$.
:::

If the degree of vertex $i$ at time $t$ is denoted by $k_i(t)$, then preferential attachment takes the form

$$
\Pi_i(t) = \frac{k_i(t)}{\sum_j k_j(t)}.
$$

This means that highly connected vertices are more likely to receive new links. In informal language, “the rich get richer.”

### Interpretation

Preferential attachment is not merely a mathematical trick. It encodes a plausible dynamic for many real systems.

-   a popular web page is more likely to attract new links,
-   a highly cited article is more likely to receive future citations,
-   a well-connected airport is more likely to gain new routes,
-   a socially visible person is more likely to form further connections.

So preferential attachment captures a cumulative-advantage mechanism.

::: {.callout-note title="Key result: preferential attachment rule"}
In a growing network with preferential attachment, the probability that a new vertex connects to existing vertex $i$ is

$$
\Pi_i(t) = \frac{k_i(t)}{\displaystyle\sum_j k_j(t)}.
$$

Since $\sum_j k_j(t) = 2L(t)$, this simplifies to $\Pi_i(t) = k_i(t) / 2L(t)$. A vertex with twice the degree is twice as likely to receive a new edge.
:::

### Why preferential attachment matters

If edges were added uniformly at random to existing vertices, the graph would not generate strong hub structure in the same way. Preferential attachment amplifies pre-existing degree differences.

Once a vertex gains an advantage, it becomes more likely to gain future links, which can strengthen the advantage further. This feedback is the source of hub formation.

## The Barabási--Albert model

::: {#def-barabasi-albert}
The **Barabási--Albert model** is defined by the following growth rule:

1.  begin from a connected seed graph,
2.  at each time step add one new vertex,
3.  connect the new vertex to $m$ existing vertices,
4.  choose each target vertex with probability proportional to its degree.
:::

The model therefore combines two ingredients:

-   **growth**: the number of vertices increases over time
-   **preferential attachment**: attachment probability is degree-proportional

Both ingredients are essential. Growth alone does not produce scale-free structure, and preferential choice without growth does not generate the same degree dynamics.

### Average degree in the BA model

At each time step, one new vertex and $m$ new edges are added. For large network size, the average degree tends to

$$
\langle k \rangle \approx 2m.
$$

This follows from the fact that each new edge contributes $2$ to the total degree and that the number of edges grows linearly with the number of added vertices.

So the parameter $m$ controls the density of the graph.

## Growth of a preferential-attachment network

The next figure illustrates the evolution of a Barabási--Albert network as new vertices are added.

```{r}
#| label: ba-growth-snapshots
#| echo: false
#| fig-cap: "Growth of a preferential-attachment network. Older highly connected vertices begin to dominate as the network expands."

g1 <- sample_pa(n = 12, power = 1, m = 2, directed = FALSE)
g2 <- sample_pa(n = 25, power = 1, m = 2, directed = FALSE)
g3 <- sample_pa(n = 50, power = 1, m = 2, directed = FALSE)

plot_ba_growth <- function(g, main_text) {
  deg <- degree(g)
  plot(
    g,
    layout = layout_with_fr(g),
    vertex.size = 8 + 2.5 * sqrt(deg + 1),
    vertex.color = course_cols$coral,
    vertex.frame.color = "white",
    vertex.label = NA,
    edge.color = scales::alpha(course_cols$ink, 0.20),
    edge.width = 1.1,
    main = main_text
  )
}

old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 3), mar = c(1, 1, 3, 1))

plot_ba_growth(g1, "Early stage")
plot_ba_growth(g2, "Intermediate stage")
plot_ba_growth(g3, "Later stage")

par(old_par)
```

This figure shows the central mechanism of the model:

-   early fluctuations matter,
-   some vertices acquire an advantage early,
-   that advantage is amplified over time,
-   hubs gradually emerge.

So the model does not treat all vertices symmetrically across time. Age and accumulated degree both influence network structure.

## Degree distribution in the BA model

A classical result for the Barabási--Albert model is that the degree distribution follows a power law with exponent

$$
\gamma = 3.
$$

More precisely, in the asymptotic limit,

$$
P(k) \sim Ck^{-3}.
$$

This means that the degree distribution is heavy-tailed: high-degree vertices are rare, but much more common than in models with exponentially decaying tails.

### Consequences of a heavy tail

The power-law form has several important structural implications.

1.  Most vertices have small degree.
2.  A few vertices may have very large degree.
3.  There is no narrow characteristic degree scale in the tail.
4.  Hub structure becomes a natural outcome of the dynamics.

This is what makes scale-free networks qualitatively different from Erdős--Rényi random graphs, whose degree distributions are much more concentrated.

::: {.callout-important title="Classical result: BA degree distribution"}
The Barabási--Albert model with parameter $m$ produces a power-law degree distribution with exponent $\gamma = 3$:

$$
P(k) \sim C k^{-3}, \qquad k \gg m.
$$

This result is independent of $m$. The parameter $m$ controls the density ($\langle k \rangle \approx 2m$) but not the tail exponent. Heavy tails are an inevitable consequence of the growth-plus-preferential-attachment mechanism.
:::

## Numerical exploration

### Degree distribution of a BA network

```{r}
#| label: ba-degree-distribution-data

set.seed(2040)

n_ba <- 3000
m_ba <- 2

g_ba_large <- sample_pa(n = n_ba, power = 1, m = m_ba, directed = FALSE)
deg_ba <- degree(g_ba_large)

deg_tbl <- tibble(k = deg_ba) |>
  count(k, name = "count") |>
  filter(k > 0) |>
  mutate(
    prob = count / sum(count)
  )

deg_tbl
```

```{r}
#| label: ba-degree-distribution-plot
#| fig-cap: "Empirical degree distribution of a Barabási--Albert network on log-log axes."

ggplot(deg_tbl, aes(x = k, y = prob)) +
  geom_point(color = course_cols$teal, size = 2.0, alpha = 0.8) +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    title = "Degree distribution of a preferential-attachment network",
    subtitle = "The heavy tail is visible on log-log axes",
    x = "Degree k",
    y = "P(k)"
  )
```

This plot does not by itself prove a power law, but it shows the heavy-tailed structure clearly. The tail decays much more slowly than the binomial or Poisson degree distributions seen in homogeneous random graphs.

### Cumulative degree distribution

```{r}
#| label: ba-cumulative-distribution
#| echo: false
#| fig-cap: "Empirical cumulative degree distribution of a Barabási--Albert network on log-log axes."

cum_tbl <- deg_tbl |>
  arrange(desc(k)) |>
  mutate(
    ccdf = cumsum(prob)
  ) |>
  arrange(k)

ggplot(cum_tbl, aes(x = k, y = ccdf)) +
  geom_point(color = course_cols$coral, size = 2.0, alpha = 0.8) +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    title = "Cumulative degree distribution",
    subtitle = "A cumulative representation often gives a smoother view of the tail",
    x = "Degree k",
    y = "P(K >= k)"
  )
```

The cumulative distribution is often smoother than the raw degree histogram and is therefore useful when investigating heavy-tailed behavior.

## Comparison with Erdős--Rényi graphs

A direct comparison with an Erdős--Rényi graph having approximately the same average degree helps clarify the difference in degree structure.

```{r}
#| label: ba-vs-er-comparison
#| echo: false
#| fig-cap: "Degree distributions of a Barabási--Albert graph and an Erdős--Rényi graph with approximately matched average degree."

p_match <- (2 * m_ba) / (n_ba - 1)
g_er_match <- sample_gnp(n = n_ba, p = p_match, directed = FALSE, loops = FALSE)
deg_er <- degree(g_er_match)

er_tbl <- tibble(k = deg_er) |>
  count(k, name = "count") |>
  filter(k > 0) |>
  mutate(prob = count / sum(count),
         model = "Erdos-Renyi")

ba_tbl <- tibble(k = deg_ba) |>
  count(k, name = "count") |>
  filter(k > 0) |>
  mutate(prob = count / sum(count),
         model = "Barabasi-Albert")

compare_tbl <- bind_rows(ba_tbl, er_tbl)

ggplot(compare_tbl, aes(x = k, y = prob, color = model)) +
  geom_point(size = 2.0, alpha = 0.8) +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_manual(values = c("Barabasi-Albert" = course_cols$coral, "Erdos-Renyi" = course_cols$teal)) +
  labs(
    title = "Heavy-tailed versus homogeneous degree structure",
    subtitle = "The two models have similar average degree but very different tails",
    x = "Degree k",
    y = "P(k)",
    color = NULL
  )
```

The contrast is the essential one:

-   the Erdős--Rényi graph has a narrow degree distribution,
-   the Barabási--Albert graph has a broad heavy-tailed degree distribution.

This is the mathematical signature of hub formation.

## Why scale-free structure is important

Scale-free structure matters because degree heterogeneity strongly affects network behavior.

### Paths and centralization

Hubs can act as highly efficient bridges. They reduce many shortest paths and create strong centralization of flow.

### Robustness and vulnerability

A network with hubs may be tolerant to random deletion of vertices, because most randomly chosen vertices are low-degree. But it may be highly vulnerable to targeted removal of hubs.

### Spreading processes

Infections, information, or cascades can spread differently in hub-dominated networks than in homogeneous graphs, because hubs can act as powerful transmission centers.

So the shape of the degree distribution is not merely descriptive. It has dynamical consequences.

## Core formulas to remember

::: {.callout-important title="Summary"}
For scale-free networks and the Barabási--Albert model:

1. **Preferential attachment probability:** $$\Pi_i(t) = \frac{k_i(t)}{\displaystyle\sum_j k_j(t)} = \frac{k_i(t)}{2L(t)}$$

2. **Average degree in the BA model:** $$\langle k \rangle \approx 2m$$

3. **Power-law degree distribution:** $$P(k) \sim C k^{-\gamma}, \qquad \gamma = 3 \text{ for the BA model}$$

4. **Scale-free property:** $$P(ak) = a^{-\gamma} P(k) \quad \text{for all } a > 0$$

5. **Cumulative tail:** $$P(K \geq k) \sim C' k^{-(\gamma - 1)}$$

6. **Degree distribution comparison:**

| Model | Degree distribution | Typical form |
|:------|:-------------------:|:------------:|
| $G(N,p)$ | Binomial / Poisson | narrow, peaked |
| Watts--Strogatz | Near-Poisson | narrow, peaked |
| Barabási--Albert | Power law | heavy-tailed, hub-dominated |
:::

## In-class questions

1.  Why does a broad degree distribution imply that average degree alone is insufficient to describe the graph?
2.  What is the difference between a heavy-tailed degree distribution and a narrow one?
3.  Why does preferential attachment lead naturally to hub formation?
4.  Why are growth and preferential attachment both essential in the BA model?
5.  Why does a scale-free network differ structurally from an Erdős--Rényi graph even when their average degrees are similar?

## End-of-chapter exercises

The exercises below are divided into three groups:

1.  mathematical exercises,
2.  conceptual and analytical exercises,
3.  simulation and coding exercises.

### Part I. Mathematical exercises

#### Exercise 1. Average degree in the BA model

In the Barabási--Albert model, one new vertex and $m$ new edges are added at each time step.

1.  Let $N$ be the number of vertices after many growth steps. Show that the total number of edges is approximately proportional to $mN$.
2.  Deduce that the average degree is approximately $$
    \langle k \rangle \approx 2m.
    $$
3.  Explain why this approximation becomes accurate for large $N$.

------------------------------------------------------------------------

#### Exercise 2. Preferential-attachment probability

Suppose the degree of vertex $i$ at time $t$ is $k_i(t)$.

1.  Explain why the preferential-attachment rule can be written as $$
    \Pi_i(t)=\frac{k_i(t)}{\sum_j k_j(t)}.
    $$
2.  Show that $$
    \sum_i \Pi_i(t)=1.
    $$
3.  Interpret this expression probabilistically.

------------------------------------------------------------------------

#### Exercise 3. Power-law scaling

Assume that the degree distribution satisfies

$$
P(k)\sim Ck^{-\gamma}.
$$

1.  Show that $$
    P(ak)\sim a^{-\gamma}P(k)
    $$ for every fixed $a>0$.
2.  Explain why this motivates the term *scale-free*.
3.  Discuss why such a distribution does not possess a single characteristic tail scale.

------------------------------------------------------------------------

#### Exercise 4. Cumulative power-law tail

Assume that

$$
P(k)\sim Ck^{-\gamma}
$$

with $\gamma>1$.

Show heuristically that the cumulative tail satisfies

$$
P(K\ge k)\sim C'k^{-(\gamma-1)}.
$$

Explain why cumulative distributions are useful in empirical analysis.

------------------------------------------------------------------------

#### Exercise 5. Hub interpretation

Let $k_{\max}$ denote the largest observed degree in a graph.

1.  Explain why a graph with narrow degree distribution typically has moderate $k_{\max}$.
2.  Explain why a heavy-tailed degree distribution allows much larger hub degrees.
3.  Discuss why hub formation is a consequence of broad degree heterogeneity.

------------------------------------------------------------------------

#### Exercise 6. BA exponent

A classical result for the Barabási--Albert model is

$$
P(k)\sim Ck^{-3}.
$$

Explain the meaning of this statement and discuss how it differs conceptually from the binomial or Poisson degree laws encountered in earlier chapters.

### Part II. Conceptual and analytical exercises

#### Exercise 7. Growth versus static models

Write a short analytical comparison between:

-   the Erdős--Rényi model,
-   the Watts--Strogatz model,
-   the Barabási--Albert model.

Your answer should address: - whether the model is static or growing, - whether degree distribution is narrow or broad, - whether clustering is naturally high, - whether hubs appear naturally.

------------------------------------------------------------------------

#### Exercise 8. Why preferential attachment creates hubs

Give a careful explanation of why preferential attachment amplifies degree differences over time.

Your answer should address: - cumulative advantage, - positive feedback, - the role of early vertices.

------------------------------------------------------------------------

#### Exercise 9. Scale-free versus small-world

Explain why the terms *scale-free* and *small-world* describe different properties.

Your answer should explain: - what structural quantity each concept refers to, - why a network can be scale-free without being a Watts--Strogatz network, - why a network can be small-world without having a power-law degree distribution.

------------------------------------------------------------------------

#### Exercise 10. Why average degree is insufficient

Two graphs may have the same average degree but very different structure.

Discuss this statement in the context of: - Erdős--Rényi graphs, - Barabási--Albert graphs.

Your answer should explain why average degree alone does not capture heterogeneity or hub structure.

### Part III. Simulation and coding exercises

#### Exercise 11. Simulating the BA model

Write an `R` function that generates a Barabási--Albert graph with parameters $(N,m)$ and returns:

-   the average degree,
-   the maximum degree,
-   the degree sequence,
-   the empirical degree distribution.

Test the function for several values of $N$ and $m$.

------------------------------------------------------------------------

#### Exercise 12. Log-log degree plot

For a fixed Barabási--Albert graph:

1.  compute the empirical degree distribution,
2.  display it on ordinary axes,
3.  display it on log-log axes,
4.  explain why the heavy tail is more visible on log-log coordinates.

------------------------------------------------------------------------

#### Exercise 13. Cumulative distribution

For a Barabási--Albert graph with large $N$:

1.  compute the empirical cumulative degree distribution,
2.  plot it on log-log axes,
3.  compare it with the raw empirical degree distribution,
4.  discuss which representation is smoother and why.

------------------------------------------------------------------------

#### Exercise 14. Comparison with Erdős--Rényi

Generate:

-   one Barabási--Albert graph with parameters $(N,m)$,
-   one Erdős--Rényi graph with approximately the same average degree.

Then compare:

-   average degree,
-   maximum degree,
-   degree distribution,
-   visual appearance.

Discuss why the two networks differ despite matched average degree.

------------------------------------------------------------------------

#### Exercise 15. Effect of system size

Repeat the BA simulation for several values of $N$.

Study how the following quantities change with $N$:

-   maximum degree,
-   degree heterogeneity,
-   visibility of the heavy tail.

Explain why hub structure becomes more pronounced as the network grows.

### Suggested mini-project

A strong mini-project for this chapter is the following:

> Compare the degree structure of homogeneous random graphs and preferential-attachment networks under matched average degree.

A complete submission should contain:

-   mathematical definitions,
-   simulation code,
-   at least two quantitative figures,
-   a comparison of maximum degree and degree distribution,
-   a concluding discussion of hub formation.

## Concluding remarks

Scale-free networks are characterized by strong degree heterogeneity and the emergence of hubs. Their degree distributions are fundamentally different from the narrow distributions of homogeneous random graphs.

The Barabási--Albert model explains one mechanism capable of producing such structure: growth combined with preferential attachment. This mechanism shows how large inequalities in connectivity can emerge naturally from a simple probabilistic rule.

The central lesson of the chapter is that network evolution rules shape degree structure, and that broad degree distributions lead to qualitatively different network organization.
